# Benchmarking workflows (DDA) {#sec-benchmarking}

## Refit function

Below we define the `msqrobRefit` function, which we use in this chapter and 
that will be added to the next release of msqrob2.

```{r refit_function}
msqrobRefit <- function(object, formula, i, subset, fcol, name,
                        modelColumnName = "msqrobModels", ...) {
    seti <- getWithColData(object, i)
    setj <- getWithColData(object, name)
    if (any(!subset %in% rowData(seti)[[fcol]]))
        stop("Some entries in 'subset' not found in '", fcol,
             "' (rowData of set '", i, "')")
    setjRefit <- msqrobAggregate(
        seti[rowData(seti)[[fcol]] %in% subset, ],
        formula = formula, fcol = fcol, modelColumnName = modelColumnName,
        ...
    )
    rowData(setj)[[modelColumnName]][subset] <-
        rowData(setjRefit)[[modelColumnName]][subset]
    modelsNew <- rowData(setj)[[modelColumnName]]
    hlp <- limma::squeezeVar(
        var = vapply(modelsNew, getVar, numeric(1)),
        df = vapply(modelsNew, getDF, numeric(1))
    )
    for (ii in seq_along(modelsNew)) {
        modelsNew[[ii]]@varPosterior <- as.numeric(hlp$var.post[ii])
        modelsNew[[ii]]@dfPosterior <- as.numeric(hlp$df.prior + getDF(modelsNew[[ii]]))
    }
    rowData(object[[name]])[[modelColumnName]] <- modelsNew
    object
}
```

In the previous chapters we have seen three different approaches to
model the data:

- [Approach 1](#sec-tmt_protein_model): start from a PSM-level table,
  compute the protein summaries and model the data at the protein 
  level.
- [Approach 2](#sec-tmt_ion_model): start from a PSM-level table 
  and model the ion-level data to obtain protein-level results.
- [Approach 3](#sec-basics): start from a peptide-level table, 
  compute the protein summaries and model the data at the protein
  level.

There are further possible approaches to model the data: 

- Approach 4: start from a peptide-level table and model the 
  peptide-level data to obtain protein-level results. 
- Approach 5: start from a protein-level table (generated by MaxQuant
  using maxLFQ) and model the protein-level data to obtain 
  protein-level results.

In this chapter, we will attempt to understand whether these different
approaches lead to difference in modelling performance. To
explore these differences, we will conduct a benchmarking experiment 
using the [E. coli spike-in experiment](#sec-ecoli_experiment),
containing ground truth information that will be used for an objective
comparison of the workflows. 

**Important**: the first sections of the chapter are meant for 
advanced users that are familiar with R scripting since benchmarking
requires some degree of automation. However, for novice users
interested in the key messages of the benchmarking and that want to
implement the best practices, we refer to the [take home message
section](#sec-take_home) for more accessible guidelines.

## Load packages

We load the `msqrob2` package, along with additional packages for
data manipulation and visualisation.

```{r load_libraries}
suppressPackageStartupMessages({
library("msqrob2")
library("dplyr")
library("tidyr")
library("data.table")
library("ggplot2")
library("patchwork")
})
```


We also configure the [parallelisation](#sec-parallel) framework.

```{r parallel}
library("BiocParallel")
register(SerialParam())
```

## Data

We will reuse the data by [@Shen2018-gw] as in @sec-basics. The
[data](#sec-ecoli_data) were reanalysed using MaxQuant, which
generates results at three levels: the evidence file containing the
[PSM table](#sec-psm_table), the [peptides file](#sec-peptide_table),
and the protein group file.

### Data files

We here retrieve those three data files.


```{r download_data}
library("BiocFileCache")
bfc <- BiocFileCache()
evidenceFile <- bfcrpath(bfc, "https://github.com/statOmics/msqrob2data/raw/refs/heads/main/dda/shen/evidence.zip")
peptideFile <- bfcrpath(bfc, "https://github.com/statOmics/msqrob2data/raw/refs/heads/main/dda/shen/peptides.txt")
proteinGroupsFile <- bfcrpath(bfc,"https://github.com/statOmics/msqrob2data/raw/refs/heads/main/dda/shen/proteinGroups.txt")
```

The data consists of a spike-in experiment with  E. coli lysates spiked at five
different concentrations (3%, 4.5%, 6%, 7.5% and 9% wt/wt) in a trypsin-digested human background.

```{r define_concentrations}
concentrations <- (2:6) * 1.5
names(concentrations) <- LETTERS[1:5]
concentrations
```

In this example, the annotations are extracted from the
sample names, although reporting a detailed design of experiments in a
table is seen as better practice [@Gatto2023-kk].
Since the evidence, peptides and protein-groups tables all contain the same samples, the annotation table will be shared across the MaxQuant data tables.

- Load the evidence table.

```{r read_evidence}
evidence <- fread(evidenceFile, check.names = TRUE, integer64 = "double")
```

The evidence file is a long table. 
Information on the raw files is stored in column Raw.file

```{r extract_rawfilenames}
rawfiles <- unique(evidence$Raw.file)
rawfiles
```


With QFeatures we can read data in long tables, but the annotation file than 
needs to have a column named runCol. 
Information on the spikein condition is indicated with a letter, A - E, which is on 
position 27 in the file name.
The acquisition was don in an end-over-end and alternating manner: A1, B1, C1, 
D1, E1, E2, D2, C2, B2, A2, ...
This can be seen if we sort the raw files alphabetically. 
Positions 5 and 6 in the file name seem to indicate the acquisition order. 

```{r sort_rawfilesnames}
(
  rawfiles <- sort(rawfiles)
)
```

Apparently sample A1 has been rerun. 

```{r rawfilenames_correctOrder}
(
  rawfiles <- rawfiles[c(20,1:19)]
)
```

```{r make_annotation}
(
  coldata <- data.frame(runCol = rawfiles) |> 
    mutate(Condition = substr(runCol,27,27),
           Concentration = concentrations[Condition],
           Sample = paste0(
             rep(letters[c(1:5,5:1)],2), 
             rep(1:4, each = 5))
           )
)
```

We retrieve all the E. coli protein identifiers to later identify
which proteins are known to be differentially abundant (E. coli
proteins) or constant (human) across condition.

```{r get_ecoli_proteins}
ecoli <- bfcrpath(bfc, "https://github.com/statOmics/msqrob2data/raw/refs/heads/main/dda/shen/ecoli_up000000625_7_06_2018.fasta")
ecoli <- readLines(ecoli)
ecoli <- ecoli[grepl("^>", ecoli)]
ecoli <- gsub(">sp\\|(.*)\\|.*", "\\1", ecoli)
```

### Convert to QFeatures

We combine each MaxQuant file with the annotation table into a 
[`QFeatures` object](#sec-qfeatures).

```{r evidences_qfeatures}
evidence <- readQFeatures(
    evidence, 
    colData = coldata, 
    runCol = "Raw.file", 
    quantCols = "Intensity")
```

- Convert the peptide table.

```{r peptides_qfeatures}
peptides <- data.table::fread(peptideFile, check.names = TRUE, integer64 = "double")
coldata$quantCols <- paste0("Intensity.", coldata$Sample)
peptides <- readQFeatures(
    peptides, colData = coldata, name = "peptides", 
    fnames = "Sequence"
)
```

- Load and convert the protein-groups table. The MaxQuant preprocessed and 
normalised data are stored in the columns that start with "LFQ.intensity."
So, we adjust the quantCols accordingly.

```{r protein_groups_qfeatures}
proteinGroups <- data.table::fread(
  proteinGroupsFile, 
  check.names = TRUE, 
  integer64 = "double")

coldata$quantCols <- paste0("LFQ.intensity.", coldata$Sample)

proteinGroups <- readQFeatures(
    proteinGroups, 
    colData = coldata, 
    name = "proteinGroups",
    fnames = "Protein.IDs"
)
```

## Data preprocessing

To exclude differences in data processing between the different data 
files, we will create a common data processing workflow in a
dedicated function^[Another option would be to copy-paste the workflow
code for every approach, but we refrain from doing so as this can lead
to incoherent code when the workflow needs to be changed. This is a
common malpractice that reduces code maintainability.].

### Preprocessing workflow

We will use the same `QFeatures` data preprocessing workflow as
presented in the [basic concepts chapter](#sec-basic_preprocess)
starting from the peptides table. However, we will account for the
increased complexity in filtering when starting from PSM-level tables,
where we need to exclude PSM mapping to multiple ions (as explained in
the [advanced concepts chapter](#sec-duplicated_psms)). Conversely,
the protein group table (containing LFQ normalised data) already
underwent several data processing steps that will be ignored. We also
need to provide the correct protein identifiers across data levels. We
will use the `Proteins` column for the PSM-level and the peptide-level
data, while we will use the `Protein.IDs` for the protein-group-level
data.

Here is an overview of the data processing workflow: 

1. [Encode](#sec-encode_missing) missing values with `NA`
2. Perform [log2-transformation](#sec-log2)
3. (Only for PSM-level data) Remove [duplicated 
   PSMs](#sec-duplicated_psms) for each ion and join the data from
   different runs, effectively leading to ion-level data.
4. (Only for protein-level data) Rename `Protein.IDs` to `Protein` for
   consistent protein identifier column name. 
5. [Filter features](#sec-filter) by removing failed protein inference
   and removing decoys and contaminants.
6. Filter missing values, keeping features that are observed in at 
   least 4 samples. This is the last step for the workflow starting 
   from protein group data, because maxLFQ values are already
   normalised and summarised to protein values.
7. Perform [normalisation](#sec-norm). 
8. Perform [summarisation](#sec-summarisation). The summarised values 
   will only be used for protein-level modelling while the normalised
   data prior to normalisation will be used for ion-level and
   peptide-level modelling. 

The workflow is implemented in a functions with one argument, `object`,
which is the `QFeatures` we generated, either from the evidence file, 
the peptides file or the protein-groups file. 

```{r preprocessing_workflow}
preprocessing_workflow <- function(object) {
    ## 1. Encode missing values
    i <- names(object) ## store the input set names
    object <- zeroIsNA(object, i)
    ## 2. Log transformation
    logName <- paste0(i, "_log") ## generate the log set names
    object <- logTransform(object, i, name = logName, base = 2)
    ## 3. PSM filtering (only for psm-level data)
    if (length(i) > 1) {
        for (ii in logName) {
            rowdata <- rowData(object[[ii]]) 
            rowdata$ionID <- paste0(rowdata$Sequence, rowdata$Charge) 
            rowdata$rowSums <- rowSums(assay(object[[ii]]), na.rm = TRUE)
            rowdata <- data.frame(rowdata) |>
                group_by(ionID) |>
                mutate(psmRank = rank(-rowSums))
            rowData(object[[ii]]) <- DataFrame(rowdata)    
        }
        object <- filterFeatures(object, ~ psmRank == 1, keep = TRUE)
        i <- "evidence"
        joinName <- paste0(i, "_log")
        object <- joinAssays(object, logName, joinName, "ionID")
        logName <- joinName 
    }
    ## 4. Match protein identifiers across data levels
    if (i == "proteinGroups") { 
        ## We need to match the protein IDs in the evidence/peptide
        ## table with the protiein IDs in the protein group table
        rowdata <- rbindRowData(object, names(object))
        rowdata$Proteins <- rowdata$Protein.IDs
        rowData(object) <- split(rowdata, rowdata$assay)
    }
    ## 5. Feature filtering
    object <- filterFeatures(
        object, ~ Proteins != "" & ## Remove failed protein inference
          !grepl(";", Proteins) & ## Remove protein groups
          Reverse != "+" & ## Remove decoys
          (Potential.contaminant != "+") ## Remove contaminants
    )
    ## 6. Missing value filtering
    n <- ncol(object[[logName]])
    object <- filterNA(object, i = logName, pNA = (n - 4) / n)
    ## (The steps below are not for protein-group data)
    if (i == "proteinGroups") return(object)
    ## 7. Normalisation
    normName <- paste0(i, "_norm")
    nfLog <- nfLogMedianOfRatios(object, logName)
    object <- sweep(
      object, MARGIN = 2, STATS = nfLog, i = logName, name = normName
    )
    ## 8. Summarisation
    summName <- paste0(i, "_proteins")
    aggregateFeatures(
        object, i = normName, name = summName, fcol = "Proteins",
        fun = MsCoreUtils::robustSummary
    )
}
```

### Run the workflow

Now that we have defined a common data processing function, we apply
to each input data level: the ion data, the peptide data and the
protein-group data. We then combine all the data in a single
`QFeatures` object.

```{r run_preprocessing, warning = FALSE, message = FALSE}
evidence <- preprocessing_workflow(evidence)
peptides <- preprocessing_workflow(peptides)
proteinGroups <- preprocessing_workflow(proteinGroups)
spikein <- c(evidence, peptides, proteinGroups)
```

## Data analysis workflows

Similarly to the data processing, we will use the same modelling 
workflow for each data level. For ion-level data (starting from the
evidence file) and peptide-level data, we will both model the data
before and after summarisation. In other words, we will model the data
at the ion/peptide level (as described in the [advanced
chapter](#sec-tmt_ion_model)) and at the protein level (as described in
the [basics chapter](#sec-modelling)).

### Model definition

Depending on the input level (ion, peptide or protein), different 
models are required. The simplest model is the protein model where we
account for the effect of `Condition` (spike-in amount group) using a
[fixed effect](#sec-fixed_effect). For the peptide and the ion models,
the data contains multiple peptides per proteins, and we expect that
intensities from the same peptide are more similar than intensities
from different peptides. Similarly, modelling multiple peptides per
protein implies that we have multiple intensities per sample and hence
that intensities from the same sample are more similar that
intensities from different samples. The ion and peptide models 
therefore include a random effect for ion (`ionID`) and peptide
(`Sequence`), and a random effect for sample (`Sample`). All three
model definitions are stored in a list to streamline later access.

```{r define_models}
models <- list(
    protein = ~ Condition,
    peptide = ~ Condition + (1 | Sample) + (1 | Sequence),
    ion = ~ Condition + (1 | Sample) + (1 | ionID)
)
```

We will benchmark the performance of the modelling approaches by
comparing all possible combination of 2 spike-in conditions^[There are
5 conditions, so 10 unique pairwise combinations.]. All models assess
the same comparisons of conditions, hence the same contrasts. We
therefore build a [contrast matrix](#sec-hypothesis_testing) that is
shared across all models.

```{r define_contrasts}
allContrasts <- createPairwiseContrasts(
    models$protein, colData(spikein), var = "Condition", ridge = TRUE
)
L <- makeContrast(
    allContrasts, 
    c("ridgeConditionB", "ridgeConditionC", "ridgeConditionD", "ridgeConditionE")
)
```

### Data modelling workflow

We will use the same `msqrob2` data modelling and statistical inference workflow as presented in the 
[basic concepts chapter](#sec-msqrob). However, for ion-level and peptide-level
models, we will rely on mixed models that have been introduced in the
[advanced chapter](#sec-run_model), including the
[refitting](#sec-msqrob_refit) of proteins for which there is only 1
feature (ion or peptide). We implement the workflow in a function that
we will call for all modelling approaches. It has 5 arguments:

- `object` is the preprocessed `QFeatures` object containing the data
  to model.
- `i` is the name of the set to start (see `names(spikein)`).
- `model` is a formula defining the model to estimate.
- `L` is the contrast matrix to perform hypothesis testing.
- `modelLevel` indicates whether the model should be fit at the 
  `"ion"`, `"peptide"` or `"protein"` level.

```{r setup_modelling_workflow, message = FALSE, warning = FALSE}
modelling_workflow <- function(object, i, model, L, modelLevel) {
    ## 1. Estimate the model
    if (modelLevel == "protein") {
        object <- msqrob(
            object,  i = i, formula = model,
            ridge = TRUE, 
            robust = TRUE
        )
    } else {
        ## 1a. Estimate
        object <- msqrobAggregate(
            object,  i = i, formula = model, 
            fcol = "Proteins", name = "msqrob",
            ridge = TRUE,
            robust = TRUE 
        )
        i <- "msqrob"
        ## 1b. Refit one-hit-wonders
        counts <- aggcounts(object[["msqrob"]])
        oneHitProteins <- rownames(counts)[rowMax(counts) == 1]
        object <- msqrobRefit(
          object, i = i, subset = oneHitProteins,
          formula = ~ Condition,
          fcol = "Proteins", name = "msqrob",
          robust = TRUE, ridge = TRUE
        )
    }
    ## 2. Hypothesis testing
    object <- hypothesisTest(object, i, contrast = L)
    ## 3. Collect the results
    out <- msqrobCollect(object[[i]], L, combine = TRUE)
    out
}
```

### Run the workflow

We will now run the above function for the different approaches. To do
so, we create a table containing the main modelling settings^[Note the
approach names match some of the sets in the `QFeatures` object.]:

- Approach 1 (called `evidence_proteins`): model the data from the 
  evidence file using the protein level data.
- Approach 2 (called `evidence_norm`): model the data from the 
  evidence file using the normalised ion data
- Approach 3 (called `peptides_proteins`): model the data from the 
  peptides file using the protein level data.
- Approach 4 (called `peptides_norm`): model the data from the 
  peptides file  using the normalised peptide data.
- Approach 5 (called `proteinGroups_log`): model the data from the 
  protein groups file using the protein level data.

```{r define_workflows}
(approaches <- data.frame(
    inputLevel = c("evidence_proteins", "evidence_norm", "peptides_proteins", "peptides_norm", "proteinGroups_log"),
    modelLevel = c("protein", "ion", "protein", "peptide", "protein")
))
```

For each row of this table, we retrieve the different arguments and
run our data modelling workflow. Notice how every approach is run
using the same code. Every `apply` iteration returns a table of
statistical inference results for all pairwise comparisons. We combine
all the tables in a single table for result exploration.

```{r run_modelling_workflow, warning = FALSE, message = FALSE}
results <- apply(approaches, 1, function(x) {
    out <- modelling_workflow(
        spikein, i = x[["inputLevel"]], model = models[[x[["modelLevel"]]]],
        L = L, modelLevel = x[["modelLevel"]]
    )
    out$inputLevel <- x[["inputLevel"]]
    out$modelLevel <- x[["modelLevel"]]
    out
})
results <- do.call(rbind, results)
```

We also add whether each modelled protein is a E. Coli protein (known
to be differentially abundant) or not. We also simplify the naming of
the contrasts for visualisation.

```{r add_annotation_results}
results$isEcoli <- results$feature %in% ecoli
results$contrast <- gsub("ridgeCondition", "", results$contrast)
results$contrast <- gsub("^([B-E])$", "\\1 - A", results$contrast)
```

## Performance benchmark{#sec-performance}

We now can compare the performance of the different modelling 
approaches. We will compare the approaches based on 4 objectives 
criteria: 

1. The number of fitted proteins across conditions.
2. The number of true positives and false positives at a 5% FDR 
   threshold. 
3. The sensitivity against the rate of missidentification.
4. The accuracy and pricision of the log2-fold change estimatione

### Number of fits

We compare the number of proteins that could be estimated by each 
approach, where a model that fits more proteins is preferred. Note 
that the number of fits exceeds the number of measured proteins 
because we summed the fits over all comparisons.

```{r number_fits, benchmarking_nfits, fig.cap=" "}
group_by(results, inputLevel) |> 
    summarise(n = sum(!is.na(adjPval))) |> 
    ggplot() +
    aes(x = inputLevel, y = n, fill = inputLevel) +
    geom_bar(stat = "identity")
```

We can see that the approaches that fits least proteins is when
starting from the protein-group file. However, all other four
approaches can fit the same number of proteins.

### TP and FP at 5% FDR

An approach that fits more proteins is preferred, provided that the
additional fits lead to meaningful results. Because this data set
contains ground truth information, we can assess whether the modelling
approaches correctly prioritised the proteins given the known
differential abundant proteins. We therefore benchmark the approaches
by examining the number of reported proteins that are true positive
(TP) (i.e., E. Coli proteins), and false positive (FP) (i.e., human
proteins) considering a 5% false discovery rate (FDR) threshold, which
is typically used. 

We first construct the table with TPs and FPs obtained from each data analysis 
workflow approach for each comparison.

```{r tp_fp_table, fig.height=20, fig.width=6}
tpFpTable <- group_by(results, inputLevel, contrast) |>
    filter(adjPval < 0.05) |>
    summarise("True Positives" = sum(isEcoli),
              "False Positives" = sum(!isEcoli)) |>
    pivot_longer(cols = c("True Positives", "False Positives"))
```

We then plot the table as a bar plot, facetting for every comparison. 

```{r benchmarking_tp_fp, fig.height=15, fig.width=4, fig.cap=" "}
ggplot(tpFpTable) +
    aes(x = inputLevel,
        y = value,
        fill = inputLevel) +
    geom_bar(stat = "identity") +
    facet_grid(contrast ~ name, scales = "free") +
    labs(x = "", y = "Count") +
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          legend.position = "bottom") +
    guides(fill = guide_legend(nrow = 3))## to avoid legend getting cropped 
```

The plot clearly shows that starting from MaxQuant's proteinGroups file leads to 
a suboptimal performance as the number of TP is systematically
lower compared to other approaches, while there are similar or even much 
more FPs. 

The four other approaches lead to comparable results. We could argue
that for some specific comparisons (C-B, D-C, E-D) approaches starting
from the evidence file recover slightly more TP without impact on the
number of FP. Similarly modelling at the peptide/ion level leads to
slight increase in performance compared to modelling at the protein
level, but these difference are subtle.

### TPR-FDP  curves{#sec-tpr_fdp}

Additionally, we construct true positive rate (TPR)-false discovery
proportion (FDP) plots. TPR represents the fraction of truly DA
proteins reported by the method, calculated as $TPR =
\frac{TP}{TP+FN}$, with FN false negatives (i.e., E. Coli proteins
that were not flagged as differential abundant). FDP denotes the
proportion of false positives among all proteins flagged as
differential abundant, calculated as $FDP = \frac{FP}{TP + FP}$.

So we first compute the FDP and TPR using the custom functions below:

```{r define_tpr_fdp_functions}
computeFDP <- function(pval, tp) {
    ord <- order(pval)
    fdp <- cumsum(!tp[ord]) / 1:length(tp)
    fdp[order(ord)]
}
computeTPR <- function(pval, tp, nTP = NULL) {
    if (is.null(nTP)) nTP <- sum(tp)
    ord <- order(pval)
    tpr <- cumsum(tp[ord]) / nTP
    tpr[order(ord)]
}
```

Before computing these metrics, we first remove any failed
inference^[When statistical inference fails for a protein, `msqrob2`
will fill the (adjusted) p-value and log2-fold change with missing
values. So excluding NAs will ignore any failed inference.]. This
means that we are comparing the approach based on a set of proteins
that is specific to each approach. 

We compute the TPR and FDP for each approach (given by `inputLevel`)
and each pairwise spike-in comparison (given by `contrast`).

```{r calculate_tpr_fdp}
performance <- group_by(results, inputLevel, contrast) |> 
    na.exclude() |>
    mutate(tpr = computeTPR(pval, isEcoli),
           fdp = computeFDP(pval, isEcoli)) |>
    arrange(fdp)
```

We also highlight the observed FDP at a 5\% FDR threshold. Since the
FDR represent the expected FDP, i.e. the average of the FDPs obtained
when the spike-in experiment were to be repeated an infinite number of
times, an observed FDP that is very far away from 5\% is indicative
for a workflow that provides poor FDR control.

```{r calculate_workpoints}
workPoints <- group_by(performance, inputLevel, contrast) |>
    filter(adjPval < 0.05) |>
    slice_max(adjPval) |>
    filter(!duplicated(inputLevel))
```

We can now generate the TPR-FDP curves. The best performing approach 
is characterised by the largest area under the curve. These curves
provide the performance over a range of FDP values, but we limit the 
plot to the $[0, 0.2]$ range because researchers are rarely interest
in the performance when the FDP exceeds 20\%.

```{r benchmarking_tpr_fdp_curve1, fig.width=10, fig.height=8, fig.cap=" "}
ggplot(performance) +
    aes(y = fdp,
        x = tpr,
        colour = inputLevel) +
    geom_line() +
    geom_point(data = workPoints, size = 3) +
    geom_hline(yintercept = 0.05, linetype = 2) +
    facet_wrap(~ contrast) +
    coord_flip(ylim = c(0, 0.2)) +
    theme(legend.position = "bottom")
```

The plots show that the msqrob2 workflows starting from the evidence and peptide 
file perform similarly. 
However, workflows starting from the proteinGroups file seem to performe worse 
for some comparisons and better for others. 

This is a bit counter intuitive as we noticed a lower sensitivity when starting
from the protein groups file when considering the number of TPs and FPs that 
were  returned at the nominal 5% FDR level. 

Note, that the performance curves above, however, do not provide a fair 
comparison. The workflows starting from the evidence and peptides file can 
fit more proteins. 

So, for a fair comparison, we also (1) create TPR-FDP plots when considering all 
proteins considered when starting from either the evidence, peptides or 
proteinGroups file. Similarly, we also compare the approaches based on (2) 
a common set of proteins that have been fit by all approaches. 

1. In the code for the TPR-FDP plot considering all proteins we calculate the 
TPR based on the union of spikein proteins that were picked with at least one 
workflow, i.e. we first calculate the total number of true positives (E. coli) 
proteins that were considered by at least one method. Note that we do not have 
to remove results with missing inference as these will appear at the end of the 
TPR-FDP plot. The code acts as if the p-values for these proteins are equal to 
1. 

```{r benchmarking_tpr_fdp_2, fig.width=10, fig.height=8, fig.cap=" "}
nTP <- results |> 
  filter(isEcoli) |> 
  pull(feature) |> 
  unique() |> 
  length() #1.
performance_all <- group_by(results, inputLevel, contrast) |> 
    mutate(tpr = computeTPR(pval, isEcoli,nTP = nTP),
           fdp = computeFDP(pval, isEcoli)) |>
    arrange(fdp)
workPoints <- group_by(performance_all, inputLevel, contrast) |>
    filter(adjPval < 0.05) |>
    slice_max(adjPval) |>
    filter(!duplicated(inputLevel))
ggplot(performance_all) +
    aes(y = fdp,
        x = tpr,
        colour = inputLevel) +
    geom_line() +
    geom_point(data = workPoints, size = 3) +
    geom_hline(yintercept = 0.05, linetype = 2) +
    facet_wrap(~ contrast) +
    coord_flip(ylim = c(0, 0.2)) +
    theme(legend.position = "bottom") +
    ggtitle("TPR-FDP curves using all proteins that were quantified")
```

Note, that this leads to a dramatic decrease in the performance for workflows 
starting from the protein groups file as way less proteins are considered / 
fitted. This leads to a much lower sensitivity and is in line with the results 
we previously observed when concidering the TPs and FPs returned at the 5% 
FDR-level.

2. In the code for the TPR-FDP plot considering the common proteins, we add a
filtering step where we require, for each comparison, that results are present 
for all 5 approaches, effectively focusing on the set of proteins that have been 
estimated by all approaches. 

```{r benchmarking_tpr_fdp_3, fig.width=10, fig.height=8, fig.cap=" "}
performance_common <- group_by(results, feature, contrast) |> 
  filter(!is.na(adjPval)) |> 
  filter(n() == 5) |> #2
  group_by(inputLevel, contrast) |> 
  mutate(tpr = computeTPR(pval, isEcoli),
         fdp = computeFDP(pval, isEcoli)) |>
  arrange(fdp)
workPoints <- group_by(performance_common, inputLevel, contrast) |>
    filter(adjPval < 0.05) |>
    slice_max(adjPval) |>
    filter(!duplicated(inputLevel))
ggplot(performance_common) +
    aes(y = fdp,
        x = tpr,
        colour = inputLevel) +
    geom_line() +
    geom_point(data = workPoints, size = 3) +
    geom_hline(yintercept = 0.05, linetype = 2) +
    facet_wrap(~ contrast) +
    coord_flip(ylim = c(0, 0.2)) +
    theme(legend.position = "bottom") +
    ggtitle("TPR-FDP curves using only the proteins fit by all approaches")
```

Interestingly, we again observe a much lower sensitivity for msqrob2 starting 
from the proteingroups file when only considering the proteins that were fitted 
by all workflows. Indicating that the difference in performance does not solely 
has to do with filtering, but also with differences in normalisation and/or 
summarisation.

We also note that the FDR is often not well controlled when starting from the 
proteinGroups file. 

These results are in line with the previous benchmark results, that also 
reported that starting from the proteinGroups file often leads to a severe 
backlash on performance. 

Interestingly, workflows starting from the evidence file or the peptides file
show comparable performances, although for challenging comparisons where the 
performance is generally low (C-B, D-C, E-D) we find a subtle increase in 
performance for ion/peptide level models.
Overal, we find that these workflows also lead to a good FDR control. 
We also find that they tend to be  slightly conservative for the D-C and E-D 
comparisons and slightly too liberal for the E-A and E-B comparison. 

### Fold change boxplots{#sec-benchmark_with_boxplots}

Next to correctly prioritising the differentially abundant proteins,
another object is to correctly estimate the log2-fold change between
conditions. Since every condition contains E. Coli proteins that have
been spiked in experimentally controlled amounts, we know the real log2-fold 
change between any two conditions. We will explore the model accuracy, i.e. how 
close the estimations are from the true value on average, and the model precision, 
i.e. how narrow the estimations are spread around the average estimation. 
In this data set, there are two target values. For E. Coli proteins, the  
expected value is the experimentally induced log2-fold change. For human 
proteins, the expected value is a log2-fold change of 0, as these proteins are 
experimentally known to be constant.

We explore the results using boxplots of the estimated log2-fold changes, but we 
first create a small table with the expected values.

```{r calculate_real_logFC, fig.height=15, fig.width=5}
realLogFC <- data.frame(
  logFC = t(L) %*% lm(log2(Concentration) ~ Condition, colData(spikein))$coef[-1]
)
realLogFC$contrast <- gsub("ridgeCondition","",colnames(L))
realLogFC$contrast <- gsub("^([B-E])$", "\\1 - A", realLogFC$contrast)
```

We can now create the boxplots with the estimated log2-fold changes,
adding horizontal lines with the corresponding target values.

```{r benchmarking_boxplot, fig.height=15, fig.width=5, fig.cap=" "}
ggplot(results) +
  aes(y = logFC,
      x = isEcoli,
      colour = isEcoli) +
  geom_boxplot()  +
  scale_color_manual(
    values = c("grey20", "firebrick"), name = "",
    labels = c("Human proteins", "E. coli proteins")
  ) +
  facet_grid(contrast ~ inputLevel, scales = "free") +
  geom_hline(data = realLogFC, aes(yintercept = logFC), 
             colour = "firebrick") +
  geom_hline(yintercept = 0) +
  theme(legend.position = "bottom")
```

For all approaches, the boxplots are roughly centred on the expected 
value, indicating good model accuracy. 

## Take home messages{#sec-take_home}

We found a drop in performance when starting from the 
protein-group data compared to the other approaches, especially when considering 
the same proteins (all proteins or the common proteins) suggesting that
these workflows can lead to suboptimal results. We did not find strong
differences in performance between the remaining approaches. For some
comparisons, we found a slight increase in performance when starting
from the evidence file compared to starting from the peptides file.
Similarly, for some comparisons, we found a slight increase in
performance when modelling data at the ion/peptide level compared to 
the protein level. 

Processing the data from the evidence file only leads to one 
additional step compared to processing the data from the peptides
file: we need to find a low-level feature definition that is shared
across run, which we here defined as the ion level (i.e. the
combination of the peptide sequence and its charge). This additional
step has little impact on the complexity of the workflow. 

Modelling the data at the ion/peptide level is more advanced than 
modelling at the protein level, as it requires the inclusion of random effects 
to account for the correlation structure within and between samples and
within and between ions/peptides from the same protein. Moreover, the
data modelling is performed at the ion/peptide level, but the
statistical inference results are reported at the protein level, hence, 
no direct protein summaries are available to explore and
support the statistical outcome^[Although these protein data are
indirectly generated using a summarisation approach (c.f.
`msqrobAggregate()`).]. Therefore, we leave it to the user to decide
whether the slight improvement in performance is worth the cost of a
more complex statistical analysis [@Sticker2020-rl].

Note, that in this benchmark, we also use ridge regression (`ridge = TRUE`) as 
opposed to the [Introductory chapter to msqrob2](#sec-basics). Ridge regression  
works generally well for designs with one factor with  multiple conditions. 
For this particular study, it provide much better FDR control. However, it increases 
the computational complexity considerably. 
For more complex designs with multiple factors ridge regression might be 
suboptimal because the optimal penalty might differ between factors (and 
blocking variables). Moreover, it does not work for simple two-groups designs or
designs with only one continuous predictor. So by default `ridge = FALSE` in the 
msqrob function. 

In summary, this chapter showed how to perform benchmarking on different
types of data input. Note, that the same framework could be used to
compare different search and quantification engines. Similarly, this
framework can also be applied to compare different instruments or
analytical protocols and setups. In the [next
chapter](sec-workflow_optimisation) we demonstrate how to compare the
impact of analysis steps for the same data.

In the remainder of this chapter, we provide a recap on how to perform
a proteomics analysis from the evidence file^[We build on the concepts
introduced in the [basic concepts chapter](#sec-basics) and the
[advanced concepts chapter](#sec-advanced)], either at the ion-level
or at the protein level.

### Preprocessing the evidence file{#sec-preprocess_evidence}

The first step is to read the data. Remember that we need two pieces
of data, the [sample annotation table](#sec-annotation_table) and, in
this case, the [PSM table](#sec-psm_table) obtained after reading
MaxQuant's evidence file.

Here are the first 6 lines (first 6 samples) of the sample
annotations, note we added `runCol` and `quantCol` that are required
for the [conversion to a `QFeatures` object](#sec_advanced_readqf). 


Here are the first 6 lines (first 6 PSMs) of the PSM table. The table
contains many columns, most containing information about the
identified peptide and the quality of the spectrum matching.

```{r}
evidence <- fread(evidenceFile, check.names = TRUE)
```

```{r, echo=FALSE}
knitr::kable(head(evidence))
```

Note that `Intensity` column contains the quantitative values and the
`Raw.file` column indicates in which run the sample was acquired. We
use the latter to link the sample annotation with the PSM table. We
therefore need to add `runCol` to the sample annotation that will
serve as the linker. 
We construct the sample annotation file from the PSM table. 

The data consists of a spike-in experiment with  E. coli lysates spiked at five
different concentrations (3%, 4.5%, 6%, 7.5% and 9% wt/wt) in a trypsin-digested human background.

```{r define_concentrations2}
concentrations <- (2:6) * 1.5
names(concentrations) <- LETTERS[1:5]
concentrations
```


```{r rawfilenames_2}
rawfiles <- unique(evidence$Raw.file)
rawfiles
```


Information on the spikein condition is indicated with a letter, A - E, which is on 
position 27 in the file name.
The acquisition was don in an end-over-end and alternating manner: A1, B1, C1, 
D1, E1, E2, D2, C2, B2, A2, ...
This can be seen if we sort the raw files alphabetically. 
Positions 5 and 6 in the file name seem to indicate the acquisition order. 

```{r sort_rawfilenames2}
(
  rawfiles <- sort(rawfiles)
)
```

Apparently sample A1 has been rerun. 

```{r correct_order_rawfilenames2} 
(
  rawfiles <- rawfiles[c(20,1:19)]
)
```

We can now generate the sample annotation table from the file names. 

```{r define_annotation2}
(
  coldata <- data.frame(runCol = rawfiles) |> 
    mutate(Condition = substr(runCol,27,27),
           Concentration = concentrations[Condition],
           Sample = paste0(
             rep(letters[c(1:5,5:1)],2), 
             rep(1:4, each = 5))
           )
)
```


We can then convert the tables to a `QFeatures`
object.

```{r evidence_qfeatures2}
(evidence <- readQFeatures(
    evidence, colData = coldata, runCol = "Raw.file", 
    quantCols = "Intensity"
))
```

Note that the data from every run is contained in a separate set. We
cannot yet join the sets together since we don't have a specific
feature identifier, yet^[A PSM is generated from a spectrum, which is
specific to each run and there is not unambiguous way to link a
spectrum across runs].

1. Encoding missing values as zeros.

```{r zeroisna2}
evidence <- zeroIsNA(evidence, names(evidence))
```

2. Log2 transforming

```{r logtransform2}
inputNames <- names(evidence)
logNames <- paste0(inputNames, "_log")
evidence <- logTransform(evidence, inputNames, name = logNames, base = 2)
```

3. Keeping only the most intense PSM per ion (see 
   [here](#sec-duplicated_psms) for a step-by-step explanation of the
   code). Upon this filtering every feature is unique to a ion
   identifier (peptide sequence + charge), and we hence can join sets
   using that identifier.

```{r psmrank_assay_joining}
for (i in logNames) {
  rowdata <- rowData(evidence[[i]]) 
  rowdata$ionID <- paste0(rowdata$Sequence, rowdata$Charge) 
  rowdata$value <- assay(evidence[[i]])[, 1]
  rowdata <- data.frame(rowdata) |>
    group_by(ionID) |>
    mutate(psmRank = rank(-value))
  rowData(evidence[[i]])$psmRank <- rowdata$psmRank
  rowData(evidence[[i]])$ionID <- rowdata$ionID
}
evidence <- filterFeatures(evidence, ~ psmRank == 1, keep = TRUE)
evidence <- joinAssays(evidence, logNames, "ions_log", "ionID")
```

4. Feature filtering

```{r filtering2}
evidence <- filterFeatures(
        evidence, ~ Proteins != "" & ## Remove failed protein inference
          !grepl(";", Proteins) & ## Remove protein groups
          Reverse != "+" & ## Remove decoys
          (Potential.contaminant != "+") ## Remove contaminants
)
```

5. Missing value filtering

```{r filter_missingness}
n <- ncol(evidence[["ions_log"]])
evidence <- filterNA(evidence, i = "ions_log", pNA = (n - 4) / n)
```

6. Normalisation

```{r normalisation2}
evidence <- sweep(
  evidence, 
  MARGIN = 2, 
  STATS = nfLogMedianOfRatios(evidence, "ions_log"), 
  i = "ions_log", 
  name = "ions_norm"
)
```

7. Summarisation

```{r summarisation2, warning=FALSE, message = FALSE}
evidence <- aggregateFeatures(
  evidence, i = "ions_norm", name = "proteins", fcol = "Proteins",
  fun = MsCoreUtils::robustSummary
)
```

### Modelling the preprocessed data

We can model the data either at the ion level or at the protein level.
Regardless of the modelling approach, we can readily generate the
[contrast matrix](#sec-multiple_contrasts) to assess all pairwise
comparisons between the experimental spike-in conditions.

```{r contrasts2}
allContrasts <- createPairwiseContrasts(
    ~ Condition, colData(evidence), var = "Condition", ridge = TRUE
)
L <- makeContrast(
    allContrasts, 
    c("ridgeConditionB", "ridgeConditionC", "ridgeConditionD", "ridgeConditionE")
)
```

#### Modelling at the protein-level

We first need to define the model we want to estimate, which describes
the sources of variation in the data. For the protein-level data, the
only potential source of variation identified from the experiment is
the spike-in condition of interest. We model it as a fixed effect.

```{r define_model2}
modelProtein <- ~ Condition
```

Then, we estimate the model.

```{r fit_model2, warning = FALSE, message = FALSE}
evidence <- msqrob(
  evidence,  i = "proteins", formula = modelProtein,
  ridge = TRUE, robust = TRUE
)
```

And we perform statistical inference on the estimated model 
parameters.

```{r hypothesistesting2}
evidence <- hypothesisTest(evidence, "proteins", contrast = L)
resultsProtein <- msqrobCollect(evidence[["proteins"]], L, combine = TRUE)
resultsProtein$isEcoli <- resultsProtein$feature %in% ecoli
```

Note that this workflows closely follows the workflow described in the 
[basics chapter](#sec-multiple_contrasts). We can report the result
using a volcano plot, for instance.

```{r benchmarking_volcano_proteins, fig.width=10, fig.height=6, fig.cap=" "}
resultsProtein |> 
  plotVolcano() +
    facet_wrap(~ contrast, scales = "free") +
    ggtitle("Statistical inference for all pairwise comparisons") 
```

#### Modelling at the ion-level{#sec-ion_level_modelling}

The model definition for the ion-level data is more ellaborate than
for the protein-level data. We need to account for the fact that
intensities within the same sample are more correlated than
intensities across samples. Similarly, we need to account for the fact
that intensities for the same ion will be more similar than
intensities between ions of the same proteins. On top of the condition
effect that we model as a fixed effect, we will account for these
sample and ion effects using a random effect. 

```{r define_model_ion2}
modelIon <- ~ Condition + (1 | Sample) + (1 | ionID)
```

Then, we estimate the model using `msqrobAggregate()` instead of
`msqrob()`.

```{r fit_model_ion2, warning=FALSE, message= FALSE}
evidence <- msqrobAggregate(
  evidence,  i = "ions_norm", formula = modelIon, 
  fcol = "Proteins", name = "msqrob",
  robust = TRUE, ridge = TRUE
)
```

Because some proteins are only measured by a single ion, its
corresponding sample and ion effects cannot be estimated and hence the
model for those proteins will not be estimated. We therefore refit a
simplified model for those proteins using the refitting workflow
described in the [advanced chapter](#sec-msqrob_refit).

We use the msqrobRefit function defined above, which will be added to msqrob2 in the 
next release. 

```{r refit2, warning = FALSE, message = FALSE}
counts <- aggcounts(evidence[["msqrob"]])
oneHitProteins <- rownames(counts)[rowMax(counts) == 1]
evidence <- msqrobRefit(
  evidence, i = "ions_norm", subset = oneHitProteins,
  formula = ~ Condition,
  fcol = "Proteins", name = "msqrob",
  robust = TRUE, ridge = TRUE
)
```

And we perform statistical inference on the estimated model 
parameters (same as above).

```{r hypothesistest_ion2}
evidence <- hypothesisTest(evidence, "msqrob", contrast = L)
resultsIon <- msqrobCollect(evidence[["msqrob"]], L, combine = TRUE)
resultsIon$isEcoli <- resultsIon$feature %in% ecoli
```

We can report the result using a volcano plot, for instance.

```{r benchmarking_volcano_ion, fig.width=10, fig.height=6, fig.cap=" "}
resultsIon |>  
    plotVolcano() +
    facet_wrap(~ contrast, scales = "free") +
    ggtitle("Statistical inference for all pairwise comparisons") 
```

