# Optimisation of a data analysis workflow {#sec-workflow-optimisation}

In the previous chapter we have seen how to benchmark the data
analysis performance when starting from different types of data. This
chapter presents how to assess the impact of different methods for the
same sequence of steps on the same data, effectively optimising the
workflow.

To illustrate the approach, we will again use the data set by
[@Shen2018-gw], a spike-in data set with known ground truth. The
optimisation focuses on two steps: we will explore different methods
for normalisation and summarisation.

**Important**: the first sections of the chapter are again meant for 
advanced users that are familiar with R scripting. Novice users
interested in the key messages and best practices can refer to the 
[take home message section](#sec-take_home).

## Load packages

We load the `msqrob2` package, along with additional packages for data
manipulation and visualisation.

```{r, message = FALSE, warning=FALSE}
suppressPackageStartupMessages({
library("msqrob2")
library("data.table")
library("dplyr")
library("tidyr")
library("ggplot2")
library("patchwork")
library("tidyverse") 
})
```

We also configure the [parallelisation](#sec-parallel) framework.

```{r}
library("BiocParallel")
register(SerialParam())
```

## Data

We will reuse the data by [@Shen2018-gw] as in @sec-basics and
@sec-benchmarking. The [data](#sec-ecoli_data) were reanalysed using
MaxQuant, and we will use start from the evidence file as suggested in
the [previous chapter](#sec-benchmarking).

### PSM table

We here retrieve the evidence file containig the [PSM
table](#sec-psm_table).



```{r download_data}
library("BiocFileCache")
bfc <- BiocFileCache()
evidenceFile <- bfcrpath(bfc, "https://github.com/statOmics/msqrob2data/raw/refs/heads/main/dda/shen/evidence.zip")
```

We again generate the annotation table  from the raw file names. 

The data consists of a spike-in experiment with  E. coli lysates spiked at five
different concentrations (3%, 4.5%, 6%, 7.5% and 9% wt/wt) in a trypsin-digested human background.

```{r}
concentrations <- (2:6) * 1.5
names(concentrations) <- LETTERS[1:5]
concentrations
```

Load the evidence table.

```{r}
evidence <- fread(evidenceFile, check.names = TRUE, integer64 = "double")
```

The evidence file is a long table. 
Information on the raw files is stored in column Raw.file

```{r}
rawfiles <- unique(evidence$Raw.file)
rawfiles
```


With QFeatures we can read data in long tables, but the annotation file than 
needs to have a column named runCol. 
Information on the spikein condition is indicated with a letter, A - E, which is on 
position 27 in the file name.
The acquisition was don in an end-over-end and alternating manner: A1, B1, C1, 
D1, E1, E2, D2, C2, B2, A2, ...
This can be seen if we sort the raw files alphabetically. 
Positions 5 and 6 in the file name seem to indicate the acquisition order. 

```{r}
(
  rawfiles <- sort(rawfiles)
)
```

Apparently sample A1 has been rerun. 

```{r}
(
  rawfiles <- rawfiles[c(20,1:19)]
)
```

```{r}
(
  coldata <- data.frame(runCol = rawfiles) |> 
    mutate(Condition = substr(runCol,27,27),
           Concentration = concentrations[Condition],
           Sample = paste0(
             rep(letters[c(1:5,5:1)],2), 
             rep(1:4, each = 5))
           )
)
```

We retrieve all the E. coli protein identifiers to later identify
which proteins are known to be differentially abundant (E. coli
proteins) or constant (human) across condition.

```{r}
ecoli <- bfcrpath(bfc, "https://github.com/statOmics/msqrob2data/raw/refs/heads/main/dda/shen/ecoli_up000000625_7_06_2018.fasta")
ecoli <- readLines(ecoli)
ecoli <- ecoli[grepl("^>", ecoli)]
ecoli <- gsub(">sp\\|(.*)\\|.*", "\\1", ecoli)
```



### Convert to QFeatures

We combine the evidence file with the annotation table into a 
[`QFeatures` object](#sec-qfeatures).

```{r}
(spikein <- readQFeatures(
    evidence, colData = coldata, runCol = "Raw.file", 
    quantCols = "Intensity"
))
```

We now have a `QFeatures` object with 20 sets, as many as the number
of MS runs. We cannot yet join the sets together since we don't have a
specific feature identifier and the data does not fullfill to the
modelling assumptions, yet. We will therefore preprocess the data 
first.

## Data preprocessing

We will follow the same data [preprocessing
workflow]({#sec-preprocess_evidence}) as in the previous chapter. We
will however explore different normalisation strategies and
summarisation methods.

1. Encoding missing values as zeros.

```{r}
spikein <- zeroIsNA(spikein, names(spikein))
```

2. Log2 transforming

```{r}
inputNames <- names(spikein)
logNames <- paste0(inputNames, "_log")
spikein <- logTransform(spikein, inputNames, name = logNames, base = 2)
```

3. Keeping only the most intense PSM per ion (see 
   [here](#sec-duplicated_psms) for a step-by-step explanation of the
   code). Upon this filtering every feature is unique to a ion
   identifier (peptide sequence + charge), and we hence can join sets
   using that identifier.

```{r}
for (i in logNames) {
  rowdata <- rowData(spikein[[i]]) 
  rowdata$ionID <- paste0(rowdata$Sequence, rowdata$Charge) 
  rowdata$value <- assay(spikein[[i]])[, 1]
  rowdata <- data.frame(rowdata) |>
    group_by(ionID) |>
    mutate(psmRank = rank(-value))
  rowData(spikein[[i]])$psmRank <- rowdata$psmRank
  rowData(spikein[[i]])$ionID <- rowdata$ionID
}
spikein <- filterFeatures(spikein, ~ psmRank == 1, keep = TRUE)
spikein <- joinAssays(spikein, logNames, "ions_log", "ionID")
```

4. Feature filtering

```{r}
spikein <- filterFeatures(
        spikein, ~ Proteins != "" & ## Remove failed protein inference
          !grepl(";", Proteins) & ## Remove protein groups
          Reverse != "+" & ## Remove decoys
          (Potential.contaminant != "+") ## Remove contaminants
)
```

5. Missing value filtering

```{r}
n <- ncol(spikein[["ions_log"]])
spikein <- filterNA(spikein, i = "ions_log", pNA = (n - 4) / n)
```

### Explore normalisation methods

[Remember](#sec-norm) that normalisation aims to correct for
systematic fluctuations across samples, as illustrated by plotting the
intensity distribution for each sample. The method applies the
following operation on each sample $i$ across all PSMs $p$:

$$
y_{ip}^{\text{norm}} = y_{ip} - \hat{\mu}_i
$$

with $\mathbf{y}$ the log intensities and \hat{\mu}_i$ the norm
factor.

We previously showed that the [Median of Ratios](#sec-norm)
(popularized by DESeq2) method could (partly) correct for these
systematic fluctuations. We abbreviate this method as the MedianOfRatios method.
The norm factor used by MedianOfRatios is estimated as follows:

$$
\hat{\mu}^\text{MedianOfRatios}_i = \text{median}(y_{i\cdot} - \frac{1}{n}\sum\limits_{i = 1}^ny_{i\cdot})
$$ 

We create a new set called `ions_norm_MedianOfRatios` that contains the
values normalised by MedianOfRatios.

```{r}
spikein <- sweep(
  spikein, MARGIN = 2, STATS = nfLogMedianOfRatios(spikein,"ions_log"), 
  i = "ions_log", name = "ions_norm_MedianOfRatios"
)
```

However, other methods exist, median centering being the most popular.
Median centering normalisation subtracts the median intensity of each
sample from all its measurements. A key assumption for this method is
that the majority of proteins are not differentially abundant. We
abbreviate this method as the MedianCentering method. The norm factor used by MedianOfRatios
is estimated as follows:

$$
\hat{\mu}^\text{MedianCentering}_i = \text{median}(y_{i\cdot})
$$

We create a new set called `ions_norm_MedianCentering` that contains the
values normalised by Median Centering using normalisation factors.

```{r, message =FALSE, warning=FALSE}
spikein <-  sweep(
  spikein, MARGIN = 2, STATS = nfLogMedian(spikein,"ions_log"), 
  i = "ions_log", 
  name = "ions_norm_MedianCentering"
)
```

We visually explore the impact of the normalisation methods on the
intensity distribution for each sample.

```{r wf_optimisation_after_norm, fig.cap=" "}
longForm(
  spikein[, , c("ions_norm_MedianOfRatios", "ions_norm_MedianCentering")],
  colvars = c("Concentration", "Condition")
) |>
  data.frame() |> 
  mutate(normalisation = sub("ions_norm_", "", assay)) |>
  ggplot() +
  aes(x = value,
      colour = Condition,
      group = colname) +
  geom_density() +
  facet_grid(~ normalisation, scales = "free") +
  labs(title = "Intensity distribution for all samples",
       subtitle = "After normalisation")
```

Both the median centering and median or ratio normalisation achieve
similar alignment of the intensity distributions. Although they appear
visually similar, we will see later that they can lead to different
results in statistical inference. 

Notice that the `QFeatures` object contains two diverging assays, one
for each normalisation method.

```{r wf_qf_plot_norm, fig.cap=" "}
plot(spikein, interactive = TRUE)
```


### Explore summarisation methods

With the data normalised, the next step is to
[summarise](#sec-summarisation) ion-level intensities to obtain
protein-level expression values. In this chapter, we will compare 
three summarisation methods:

- **Robust summarisation** [@Sticker2020-rl], as we described in a 
  [previous chapter](#sec-summarisation). 
  
- **Median summarisation**, which calculates the median of the 
  log-transformed intensities of all peptides belonging to a protein 
  for each sample separately. While robust to outliers, it can be 
  heavily influenced by inconsistent peptide identification across 
  samples.
  
- **Median polish summarisation**, is a simple and robust procedure proposed 
by the statistician John Tukey to fit an additive model for data in a two-way 
layout table (usually, results from a factorial experiment) of the form 
row effect + column effect + overall median. Median polish utilizes the medians 
obtained from the rows and the columns of a two-way table to iteratively 
calculate the row effect and column effect on the data. The results are robust 
to outliers, as the iterative procedure uses the medians rather than the means. 
Here, the row effects are the feature effects and the column effects are the 
sample/run effects. Particularly, following model is considered: 
$ y_{ip} = \beta_p^\text{pre} + \beta_i^\text{samp} + \epsilon_{ip}$, where 
$y_{ip}$ is the log-normalised feature intensity (precursor, PSM or peptide)
belonging to protein $P$ in sample $i$. $\beta_p^\text{pep}$ is the average 
effect of feature $p$, which account for the fact that different features yield 
different baseline intensities. $\beta_i^\text{samp}$ is the average effect of 
sample $i$. In other words, it provides the estimated log2-transformed and 
normalised intensity of protein $P$ in sample $i$ corrected for the feature 
effect, which will be used as the summarised protein value.
$\epsilon_{ip}$ is the residual effect that cannot be explained by the
average sample and precursor effects.
  
- **maxLFQ summarisation**, maxLFQ summarise the precursor-level data into protein intensities. 
It first calculates all pairwise log ratio's between samples only using their shared precursors.
Particularly, it uses the median of the log ratio's between the shared precursors s, i.e. 
$$r_{ij} = median(y_{sj}-y_{si})$$. Hence, it first eliminates peptide effects. 
It then estimates the summaries by solving $\sum_i\sum_j(y^{prot}_j - y^{prot}_i - r_ij)^2$ and 

The robust, median and median polish summarisation methods are available in 
`MsCoreUtils` and maxLFQ is implemented in the `maxLFQ` function of the `iq` 
package. All methods can be integrated with `QFeatures`.

```{r}
summMethods <- c(
  robustSummary = MsCoreUtils::robustSummary, 
  colMedians = colMedians, 
  medianPolish = MsCoreUtils::medianPolish,
  maxLFQ = function(X, ...) iq::maxLFQ(X)$estimate
)
```

We will therefore create a loop that iterates over the normalised data
sets and over the summarisation methods.

```{r, warning=FALSE}
for (i in c("ions_norm_MedianOfRatios", "ions_norm_MedianCentering")) {
  for (j in names(summMethods)) {
    newSet <- paste0("proteins_", sub("ions_norm_", "", i), "_", j)
    spikein <- aggregateFeatures(
      spikein, i = i, 
      name = newSet,
      fcol = "Proteins", 
      fun = summMethods[[j]],
      na.rm = TRUE
    ) 
  }
}
```

The data processing is now complete. The `QFeatures` object now
contains a workflow of assays, from input PSMs to normalised and
summarised proteins. The data set contains 2 sets obtained using 2
normalisation methods, and each normalised set has been summarised
with four different methods.

```{r wf_qf_plot_summarisation, fig.cap=" "}
plot(spikein, interactive = TRUE)
```

## Data modelling

The preprocessed data for all 6 workflows^[There are 2 normalisaion
methods * 3 summarisation methods.] can now be modelled to assess our
research hypothesis using the same approach, as described in the
[basics chapter](#sec-modelling)^[Note that while we changed the
preprocessing workflow, the sources of variation and the model to
estimate remain the same as in @sec-basics.]. Remember,
the central research question is to "prioritise proteins that have
been spiked in across the conditions." Since we know the ground truth,
we can assess how accurately the model estimates the fold changes
between spike-in conditions, and hence which preprocessing workflow
leads to the best results.

We loop over the different assays to fit the model for the different
preprocessed sets^[Those sets that start with `proteins_`.]. Recall
that the model estimation results are stored in the `rowData` of each
set.

```{r message = FALSE, warning = FALSE}
proteinSets <- grep("proteins_", names(spikein), value = TRUE)
for (i in proteinSets) {
  spikein <- msqrob(
    spikein,  i = i, formula = ~ Condition, 
    ridge = TRUE, 
    robust = TRUE,
    overwrite = TRUE
  )
}
```

We enable [M-estimation](#sec-robust) (`robust = TRUE`) for improved
robustness against outliers and [ridge regression](#sec-ridge) (`ridge
= TRUE`) to stabilise parameter estimates^[In our hands ridge regression works 
well for designs with one factor with multiple conditions. However, it increases 
the computational complexity considerably. For more complex designs with 
multiple factors ridge regression might be suboptimal because the optimal 
penalty might differ between factors (and blocking variables) so by default 
`ridge = FALSE` in the msqrob function]. 

We can create our [contrasts of interest](#sec-inference) and perform
[hypothesis testing](#sec-hypothesis_testing).

```{r}
allContrasts <- createPairwiseContrasts(
    ~ Condition, colData(spikein), var = "Condition" , ridge = TRUE
)
L <- makeContrast(
    allContrasts, 
    c("ridgeConditionB", "ridgeConditionC", "ridgeConditionD", "ridgeConditionE")
    )

inferences <- lapply(proteinSets, function(i) {
  spikein <- hypothesisTest(spikein, i, contrast = L)
  inference <- msqrobCollect(spikein[[i]], L)
  inference$method <- i
  inference
})
inferences <- do.call(rbind, inferences)
```

We also add the information whether a protein is differentially
abundant or not, since all E. coli proteins are known to be spiked in
different concentrations.

```{r}
inferences <- mutate(
  inferences,
  isEcoli = feature %in% ecoli,
  normalisation = sub("(.*)_(.*)_(.*)", "\\2", method),
  summarisation = sub("(.*)_(.*)_(.*)", "\\3", method),
  contrast = gsub("ridgeCondition", "", contrast),
  contrast = gsub("^([B-E])$", "\\1 - A", contrast),
)
```


## Performance Evaluation

Since we have ground truth information (i.e., we know which proteins
are from E. coli and should be differentially abundant), we can
objectively evaluate the performance of each workflow. Similarly to
the [previous chapter](#sec-performance), we will assess the number of
true positives (TP, that are correctly identified E. coli proteins)
and false positives (FP, that are human proteins incorrectly
identified as significant), the false discovery proportion (FDP), the
accuracy of the estimated log2 fold changes (LFC), and finally, the
overall performance using True Positive Rate (TPR) versus False
Discovery Proportion (FDP) curves.

Before delving into the performance plot, we will first create a
colour scheme in order to compare the different workflows. We will
assign a colour to each combination of normalisation and summarisation
method.
We also remove the prefix "ridgeCondition" in the contrasts.

```{r}
inferences <- inferences |> 
  mutate(
  workflow = paste0(normalisation, "_", summarisation),
  contrast = gsub(pattern = "ridgeCondition", "", contrast)
)
colours <- c(
  MedianCentering_colMedians = "#fcba03",
  MedianCentering_maxLFQ ="#c29310",
  MedianCentering_medianPolish = "#7a5a0a",
  MedianCentering_robustSummary = "#423512",
  MedianOfRatios_colMedians = "#5bb4fc",
  MedianOfRatios_maxLFQ = "#0b69b5",
  MedianOfRatios_medianPolish = "#084a85",
  MedianOfRatios_robustSummary = "#0f2a40"
)
```


### TP and FP at 5% FDR

We will first construct the table with TPs and FPs obtained
from each data modelling approach for each comparison.

```{r}
tpFpTable <- inferences |>
  na.omit() |>
  mutate(TP = adjPval < 0.05 & isEcoli,
         FP = adjPval < 0.05 & !isEcoli) |>
  group_by(normalisation, workflow, contrast) |>
  summarise("TruePositives" = sum(TP),
            "FalsePositives" = sum(FP)) |>
  mutate(FDP = FalsePositives / (TruePositives + FalsePositives)) |>
      pivot_longer(cols = c("TruePositives", "FalsePositives"),
                   names_to = "metric", values_to = "count")
```

We then plot the table as a bar plot, facetting for every comparison. 

```{r wf_tp_fp, fig.width=6, fig.height=15, fig.cap=" "}
ggplot(tpFpTable) +
  aes(
    x = normalisation, y = count,
    fill = workflow
  ) +
  geom_bar(
    stat = "identity", position = position_dodge(preserve = "single"),
    color = "black"
  ) +
  facet_grid(contrast ~ metric, scales = "free") +
  scale_fill_manual(values = colours) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

We see that, while it has minimal impact on the TPs, the median centering 
normalisation approach increases the number of FPs for some comparisons 
(E-A and E-B). We therefore suggest to use the Median of Ratios normalisation 
method.  

More importantly, we see that summarising using the column median leads to 
poor performance (irrespective of the normalisation method used) as these 
workflow identify less TPs.
This can be easily explained: the summarisation step in this workflow does 
not account for the fact that different PSMs are often observed for a protein 
in the different samples and that their corresponding intensities largely 
fluctuate according to the differences in the PSM characteristics. 
The median polish, maxLFQ and the robust summary summarisation methods show 
similar performance for most comparisons.  
`maxLFQ` from the `iq` package has a slightly higher sensitivity for some 
comparisons and slightly more false positives in others. 
Note, however, that median polish is computationally much less demanding and is 
recommended for designs with many samples. 


### TPR-FDP curves

We generate the [TPR-FDP curves](#sec-tpr_fdp) to assess the
performance of the different workflows to prioritise differentially
abundant proteins. Again, these curves are built using the ground
truth information about which proteins are differentially abundant
(spiked in) and which proteins are constant across samples. We create
two functions to compute the TPR and the FDP.

```{r tpr_fdp_function}
computeFDP <- function(pval, tp) {
    ord <- order(pval)
    fdp <- cumsum(!tp[ord]) / 1:length(tp)
    fdp[order(ord)]
}
computeTPR <- function(pval, tp, nTP = NULL) {
    if (is.null(nTP)) nTP <- sum(tp)
    ord <- order(pval)
    tpr <- cumsum(tp[ord]) / nTP
    tpr[order(ord)]
}
```

We apply these functions and compute the corresponding metric using the
statistical inference results and the ground truth information.

```{r, echo= FALSE}
performance <- group_by(inferences, summarisation, normalisation, contrast) |> 
    mutate(tpr = computeTPR(pval, isEcoli),
           fdp = computeFDP(pval, isEcoli)) |>
    arrange(fdp)
```

We also highlight the observed FDP at a 5\% FDR threshold.

```{r}
workPoints <- group_by(performance, summarisation, normalisation, contrast) |>
    filter(adjPval < 0.05) |>
    slice_max(adjPval)
```

We can now generate the plot^[Note that the code to build the TPR-FDP
curves is identical to the workflow in the previous chapter on data
benchmarking.].

```{r wf_fdr_fdp, fig.height=15, fig.width=5, fig.cap=" "}
ggplot(performance) +
    aes(
        y = fdp,
        x = tpr,
        colour = workflow
    ) +
    geom_line() +
    geom_point(data = workPoints, size = 3) +
    geom_hline(yintercept = 0.05, linetype = 2) +
    facet_grid(contrast ~ .) +
    coord_flip(ylim = c(0, 0.2)) +
    scale_colour_manual(values = colours)
```

Again, TPR-FDP curves clearly indicate a suboptimal performance (lower 
sensitivity at the same false discovery proportion) of median summarisation.
  
We also observe a benefit of the Median Of Ratios normalisation, which seems to 
improve the FDR control accross the board^[the dots indicating the sensitivity 
and FDP when using a 5% FDR cut-off are much closer to the empirical 5% FDP] as 
well as the sensitivity in contrasts involving the higher spike-in conditions. 

This can be explained by the difference in composition between the samples that 
is inherent to spike-in design. Indeed, median normalisation only accounts for 
differences in loading by considering within sample information only, whereas 
Median of Ratios normalisation uses both within and between sample information. 

### Fold change boxplots

Finally, we estimate the accuracy and precision of the log2-fold changes using 
[boxplots](#sec-benchmark_with_boxplots). We first create a table with the 
ground truth information from the spike-in experiment.

```{r}
realLogFC <- data.frame(
  expectedLogFC = t(L) %*% lm(log2(Concentration) ~ Condition, colData(spikein))$coef[-1]
)
realLogFC$contrast <- gsub("ridgeCondition","",colnames(L))
realLogFC$contrast <- gsub("^([B-E])$", "\\1 - A", realLogFC$contrast)
```

We now plot the estimated log2-fold changes, focusing on the 
differentially abundant proteins (E. Coli proteins). 
The dashed line represents the true log2-fold change for each comparison. 

```{r wf_boxplot, fig.height=15, fig.width=5, fig.cap=" "}
filter(inferences, isEcoli) |> 
  ggplot() +
  aes(
      y = logFC,
      x = normalisation,
      colour = workflow
  ) +
  geom_boxplot()  +
  facet_grid(contrast ~ ., scales = "free") +
  geom_hline(data = realLogFC, aes(yintercept = expectedLogFC), 
             linetype = "dashed") +
  scale_colour_manual(values = colours) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

The plots show that median summarisation produces biased fold change estimates, 
again because it does not account for the differences in the characteristics of 
the PSMs that are observed in the different samples. 

## Conclusion

In this chapter, we objectively compared different normalisation and
summarisation strategies using a spike-in dataset with known ground
truth.

1.  **Normalisation is critical**: The performance evaluation clearly
shows that the choice of normalisation method has a significant impact
on the results. Workflows using the Median of Ratios normalisation
consistently outperformed those using simple median centering. They
yielded more true positives, better control of the false discovery
proportion, and more accurate log2-fold change estimates, particularly
for smaller fold changes. This because it not only accounts for differences in 
loading between samples but also for differences in the 
compositionality. 

2. **Column median summarisation is suboptimal**: we found that 
summarising peptide data to protein intensities using the column
(sample) median leads to reduced sensitivity (low True Positive Rate)
and specificity (low False Discovery Proportion), but also to lower
accuracy (i.e. the estimated log2-fold changes are further from the
true value on average) and precision (the estimations are more spread
around the average estimation).

3. **MaxLFQ, median polish and robust summary provide comparable performance**:
the TPR-FDP curves show that the pipelines incorporating robust summarisation 
and median polish are consistently closer to the ideal top-left corner, with 
only subtle differences between the two approaches. This indicates that 
explicitly modelling peptide-specific effects and down-weighting outliers leads 
to more reliable protein-level quantification and inference. Note, that median 
polish summarisation is computationally less demanding and that we could not 
find a benefit in performance for robust summarisation that justifies its 
additional computational overhead.

Based on this comprehensive evaluation, the recommended pipeline for
this type of dataset is the **Median of Ratios normalisation**
followed by **median polish summarisation**. This combination provided
the best performance in terms of computational complexity, accuracy, sensitivity 
and FDR control, demonstrating the strength of `msqrob2` in identifying truly 
differentially abundant proteins while controlling for false discoveries.

Note, that these considerations are based on this specific spike-in study and 
that this can slightly changes for other datasets.  
However, we have seen over a wide range of examples that

- median summarisation  generally performs poorly, 
- median normalisation performs quite well in datasets with samples that have
similar complexity. 
- in our hands, median normalisation never seems to drastically outperforms 
Median of Ratios normalisation, which seems a safe option across the board. 
- ridge regression (`ridge = TRUE`) works well for designs with one factor with 
multiple conditions. For this particular study, it provide much better FDR 
control. However, it increases the computational complexity considerably. 
For more complex designs with multiple factors ridge regression might be 
suboptimal because the optimal penalty might differ between factors (and 
blocking variables) so by default `ridge = FALSE` in the msqrob function.    

