# msqrob2PTM: DIA-NN case study - Unpaired phospho-enriched and non-enriched samples {#sec-msqrob2PTM-unpaired}

## Background 

- Mass spectrometry–based (MS) proteomics allows the identification and 
  quantification of a myriad of posttranslational modifications (PTMs)
- This reveals additional complexity and diversity of the proteome. 
- Indeed, PTMs greatly extend the number of different forms of a protein, 
  that is, proteoforms, that can be found. 
- More importantly, these PTMs can impact protein functions 
- They are linked to a variety of diseases and developmental disorders. 
- Aberrant PTM status can cause a number of detrimental effects ranging from the 
  alteration of protein folding to the dysregulation of cell signaling. 
- It is thus of great importance to study these PTMs in detail, not only through 
  their correct identification but also by their correct quantification and 
  subsequent statistical analysis.
  
### Vignette 

In this vignette we start from quantified precursor level data to infer

- Differential abundance (DA) of precursors carrying post-translation-modifications, 
- DA at the protein level
- Differential usage of precursors, i.e. 
  DA of precursors upon correction for DA at protein level
- Differential usage of PTMs, i.e. upon summarising the usages for all 
  precursors that carry the same PTM at a specific residue (position in 
  the protein), 
- Specifically we will use an unpaired design where the enriched and 
  non-enriched runs originate from different bio-repeats

using the msqrob2PTM workflow in the bioconductor package [msqrob2](https://www.bioconductor.org/packages/release/bioc/html/msqrob2.html).

You can read more about 
- `msqrob2` on [https://statomics.github.io/msqrob2book/](https://statomics.github.io/msqrob2book/) 
- `msqrob2PTM` in msqrob2PTM: Differential Abundance and Differential Usage 
  Analysis of MS-Based Proteomics Data at the Posttranslational Modification and 
  Peptidoform Level. N. Demeulemeester, M. Gébelin, L. Caldi Gomes, C. Carapito, 
  L. Martens and L. Clement. Molecular & Cellular Proteomics, 2023; 23. 
  [https://doi.org/10.1016/j.mcpro.2023.100708](https://doi.org/10.1016/j.mcpro.2023.100708)

### Experiment

Mutations in tRNAs can lead to mis-incorporation of an amino acid into a growing 
polypeptide chain that differs from what is specified by the mRNA in a process 
known as mistranslation. 
To better understand the impact of mistranslating tRNA variants, Berg et al. 
profiled the proteome and phosphoproteome of yeast expressing three different 
mistranslating tRNAs. 
The first is a proline tRNA with a G3:U70 base pair in its acceptor stem that 
mis-incorporates alanine at proline codons (Pro -> Ala; Hoffman et al. 2017). 
The other two are serine tRNAs with either a UGG proline or UCU arginine 
anticodon which mis-incorporate serine at proline (Pro -> Ser) or serine at 
arginine codons (Arg -> Ser), respectively (Berg et al. 2017, 2019b). 

Six replicates of each strain were grown. From each replicate an non-enriched (whole proteome)
and a phospho-peptide-enriched sample was prepared and quantified using a data
independent acquisition workflow.

The data for the non-enriched and phospho-enriched samples are deposited to 
PRIDE, dataset identifiers PXD068388 and PXD068392). 

We will take a subset of the data to illustrate an unpaired analysis, i.e. we 
will use the phospho-data from three random samples and the non-enriched data 
for the remaining samples. 

Summary: 

- Mutations in tRNAs can lead to mis-incorporation of amino acids during protein 
  translation.
- 4 strains: WT, 3 strains with mutations leading to miss-incorporations 
  (Pro -> Ala, Pro -> Ser and Arg -> Ser).
- 6 biological repeats per strain
- Here we use the non-enriched MS runs from three random bio-repeats and 
  phospho-enriched MS runs from the other three bio-repeats so as to have 
  unpaired samples. 

M. Berg, A. Chang, R. Rodriguez-Mias, J. Villén, Mistranslating tRNA variants 
impact the proteome and phosphoproteome of Saccharomyces cerevisiae, 
G3 Genes|Genomes|Genetics, Volume 16, Issue 2, February 2026, jkaf284, 
[https://doi.org/10.1093/g3journal/jkaf284](https://doi.org/10.1093/g3journal/jkaf284)

## Preamble

### Load packages

We load the `msqrob2` package, along with additional packages for
data manipulation and visualisation.

```{r load_libraries}
suppressPackageStartupMessages({
library("QFeatures")  
library("dplyr") 
library("tidyr")
library("ggplot2")
library("msqrob2")    
library("stringr")
library("ExploreModelMatrix")
library("MsCoreUtils") 
library("matrixStats")
library("patchwork")
library("kableExtra")
library("ComplexHeatmap")
library("purrr")
library("tibble")
library("scater")
library("BiocFileCache")
})
```

## Data

### Precursor table

- We load the output from DIA-NN parquet files. 
- Can be file path to local file or url to file that lives on the web.
- Here we use BiocFileCache so that we do not need to download the files again each 
time that the script is knit. 


```{r data_path}
bfc <- BiocFileCache()
precursorFile <- bfcrpath(bfc,"https://zenodo.org/records/20414816/files/WholeProteome_DIANNreport.parquet?download=1")
precursorFilePTM <- bfcrpath(bfc,"https://zenodo.org/records/20414816/files/Phosphoproteome_DIANNreport.parquet?download=1")
```


We can import the report.parquet file using the `read_parquet` function from the 
`arrow` package. 

```{r data_import}
precursors <- arrow::read_parquet(precursorFile) # function from the arrow package
precursorsPTM <- arrow::read_parquet(precursorFilePTM) # function from the arrow package
#precursors <- data.table::fread(precursorFile) # For older versions the results are stored as tsv files. 
```

We subset the precursor files to reduce the memory footprint and only keep the 
variables that we need in the downstream analysis. 

Note that `Protein.Sites` refers to the exact modified residue position in the full 
protein sequence, which will be useful when assessing PTM-level quantification. 

Note, that we also filter out the strain with the ProSer variant here. 

```{r data_subset}
precursors <- precursors |> 
  filter(!grepl("ProSer",Run)) |>
  select(
         Run, 
         Precursor.Id, 
         Modified.Sequence, 
         Stripped.Sequence, 
         Precursor.Charge, 
         Protein.Group, 
         Protein.Names,
         Protein.Ids,
         Protein.Sites, #exact modified residue position in the full protein sequence
         Genes,
         Precursor.Quantity,
         Precursor.Normalised,
         Normalisation.Factor,
         Ms1.Area,
         Ms1.Normalised,
         PG.MaxLFQ,
         Q.Value, 
         Lib.Q.Value,
         PG.Q.Value,
         Lib.PG.Q.Value,
         Proteotypic,
         Decoy, # Not available in older versions of DIA-NN
         RT)
precursorsPTM <- precursorsPTM |> 
  filter(!grepl("ProSer",Run)) |>
  select(
         Run, 
         Precursor.Id, 
         Modified.Sequence, 
         Stripped.Sequence, 
         Precursor.Charge, 
         Protein.Group, 
         Protein.Names,
         Protein.Ids,
         Protein.Sites, #exact modified residue position in the full protein sequence
         Genes,
         Precursor.Quantity,
         Precursor.Normalised,
         Normalisation.Factor,
         Ms1.Area,
         Ms1.Normalised,
         PG.MaxLFQ,
         Q.Value, 
         Lib.Q.Value,
         PG.Q.Value,
         Lib.PG.Q.Value,
         Proteotypic,
         Decoy, # Not available in older versions of DIA-NN
         RT)
```

Garbage collection to free space 

```{r}
gc(); gc()
```

### Subset phospho

In DIA-NN phosphorylation is indicated with (UniMod:21) in the sequence stored in 
variable `Modified.Sequence`. This variable contains the precursor sequence 
along with its modifications. 

Here, we only keep precursors that are phosphorylated. 

```{r}
precursorsPTM <- precursorsPTM |>
  filter(grepl(pattern = "\\(UniMod:21\\)", Modified.Sequence))
```

Garbage collection to free space 

```{r}
gc(); gc()
```

### Quick check for imputation

Quick check on distribution of precursors MS2 intensities.

```{r quick check}
precursors |> 
  filter(is.finite(Precursor.Quantity) & Precursor.Quantity > 0) |>
  ggplot(aes(x = log2(Precursor.Quantity))) +
  geom_density() +
  theme_minimal() +
  labs(subtitle = "Abundances (non-enriched)") + precursorsPTM |> 
  filter(is.finite(Precursor.Quantity) & Precursor.Quantity > 0) |>
  ggplot(aes(x = log2(Precursor.Quantity))) +
  geom_density() +
  theme_minimal() +
  labs(subtitle = "Abundances (phospho-enriched)")
```


Seems no imputation has been done. 


```{r}
#Garbage collection to free space 
gc(); gc()
```

### Subset the data to make it unpaired

We select at random 3 biorepeats for phospho-enriched and 3 for non-enriched 
samples.

```{r}
set.seed(1245)
sampleNames <- paste0("_0",1:6,"_")
peSamp <- sample(sampleNames,3) |> sort()
neSamp <- sampleNames[!sampleNames %in% peSamp]
```

Subset the data

```{r}
precursors <- precursors |> filter(grepl(paste(neSamp,collapse="|"), Run))
precursorsPTM <- precursorsPTM |> filter(grepl(paste(peSamp,collapse="|"), Run))
```

```{r}
#Garbage collection to free space 
gc(); gc()
```

### Sample annotation tables

The [sample annotation table](#sec-annotation_table)) is not available
and can be generated from the run labels, as the researchers included information on the design in the filenames.

1. Extract unique Run labels
2. Split Run label into variables using the "_" separator
3. Rename the variable Run to runCol (needed for `readQFeatures` function)
4. Add sampleId
5. Sort annotation file according to sampleId

```{r metadata_non_enriched}
annot <- precursors |> 
  dplyr::distinct(Run) |>  ## 1.
  separate(Run,
           into=c("run","label2","strain","rep","type","acquisition"),
           sep="_",
           remove=FALSE) |>   ## 2.
  rename(runCol = Run) |> ## 3.
  mutate(sampleId = paste0(strain,rep)) |> #4.
  arrange(sampleId) #5.
annot
```

We do the same for the phospho data.

```{r meta_data_phospho}
annotPTM <- precursorsPTM |> 
   dplyr::distinct(Run) |>  ## 1.
  separate(Run,
           into=c("label1","label2","strain","rep","type","acquisition"),
           sep="_",
           remove=FALSE) |>   ## 2.
  rename(runCol = Run) |> ## 3.
  mutate(sampleId = paste0(strain,rep)) |> #4.
  arrange(sampleId) #5.
annotPTM
```

```{r}
#Garbage collection to free space 
gc(); gc()
```

### Convert to QFeatures

First, recall that the precursor table is file in long format. 
Every quantitative column in the precursor table contains
information for multiple runs. Therefore, the function split the table
based on the run identifier, given by the `runCol` argument (for
DIA-NN, that identifier is contained in `Run`). 
So, the `QFeatures` object after import will contain as many sets as there are
runs. 
Next, the function links the annotation table with the PSM data.
To achieve this, the annotation table must contain a `runCol` column
that provides the run identifier in which each sample has been
acquired, and this information will be used to match the identifiers
in the `Run` column of the precursor table. 

Here, we will use the `Precursor.Quantity` column as quantification input. 

1. We first combine the precursor and precursorPTM table in a list of 
QFeatures objects, 
2. which we subsequently reduce in one QFeatures object. Note that in 
QFeatures, the function `c()` is defined to merge QFeatures objects, stacking 
all assays together into a single object.

This allows us to store the data of the phospho-enriched and the non-enriched 
samples in the same QFeatures `object`.

```{r import_qfeatures}
(qf <- list(
  readQFeatures(assayData = precursors,  
                              colData = annot,
                              quantCols = "Precursor.Quantity",
                              runCol = "Run",
                              fnames = "Precursor.Id"),
  readQFeatures(assayData = precursorsPTM,  
                              colData = annotPTM,
                              quantCols = "Precursor.Quantity",
                              runCol = "Run",
                              fnames = "Precursor.Id") ##1.
) |> purrr::reduce(c) ##2.
)
```

Remove data tables to free space. 

```{r remove_precursor_data}
rm(precursors, precursorsPTM)
```
```{r}
#Garbage collection to free space 
gc(); gc()
```

## Data preprocessing{#sec-basic_preprocess}

Here we conduct the following processing steps

1. Replace zero's by NA.
2. Precursor filtering and assay joining
3. Log-transformation
4. Normalisation
5. Summarisation for non-enriched runs 

### Encoding missing values

We first replace any zero in the quantitative data
with an `NA`. 

```{r zero_to_na}
qf <- zeroIsNA(qf, names(qf))
```

Note that `msqrob2` can handle missing data without having to rely on
hard-to-verify imputation assumptions, which is our general recommendation. 
However, `msqrob2` does not prevent users from using imputation, which can be 
performed with `impute()` from the `QFeatures` package. 

### Precursor Filtering 

Filtering removes low-quality and unreliable precursors that would otherwise 
introduce noise and artefacts in the data. 

#### Remove questionable identifications

We apply standard filtering: 


1. q-value threshold of 0.01 for the identification of precursors (`Q.Value`) 
and protein groups (`PG.Q.Value`). If the file is processed via matching between 
runs, it is also useful to filter on Lib.Q.Value and Lib.PG.Q.Value.

2. Remove precursors that could not be mapped, i.e. when `Precursor.Id` column 
is an empty string.  

3. Filter decoys, i.e. only keep precursors for which the `Decoy` column equals 
0. (Note, that the `Decoy` column is not present in the output of older versions 
of DIA-NN).

4. Keeping only proteotypic peptides, which map uniquely to a specific protein.



```{r filtering}
qf <-  qf |>
  filterFeatures(~ Q.Value <= 0.01 & #1.
                     PG.Q.Value <= 0.01 &  #1.
                     Lib.Q.Value <= 0.01 & #1.
                     Lib.PG.Q.Value <= 0.01 & #1.
                     Precursor.Id != "" & #2.
                     Decoy == 0 & #3. (Note, that this filter is not available with previous versions of DIA-NN, as the report.tsv file did not include a Decoy column. So other strategies are needed if Decoys are in the output file.)
                     Proteotypic == 1) #4.
```


Note, that it is important that the filtering criteria are not distorting the 
distribution of the test statistics in the downstream analysis for features that 
are non-DA. 
It can be shown that filtering will not induce bias results when the filtering 
criterion is independent of test statistic. The criteria that we proposed above 
are all based on the results of the identification step, hence, they are 
independent of the downstream test statistics that will be used to prioritize DA 
proteins. 

```{r}
#Garbage collection to free space 
gc(); gc()
```

#### Assay joining

- Up to now, the data from different runs were kept in separate assays. 
- We can now join the normalised sets into an precursor set using joinAssays(). 
- Sets are joined by stacking the columns (samples) in a matrix and rows 
(features) are matched according to a row identifier, here the `Precursor.Id`.
- We do this for the non-enriched and phospho samples, which we store in assays 
with names: `precursors` and `precursorsPTM`, respectively.

```{r assay_joining}
qf <- joinAssays(
  x = qf,
  i = grep("Proteome",names(qf), value = TRUE),
  fcol = "Precursor.Id",
  name = "precursors"
  )
(qf <- joinAssays(
  x = qf,
  i = grep("Phospho",names(qf), value = TRUE),
  fcol = "Precursor.Id",
  name = "precursorsPTM"
  )
)
```

Remove the original assays to free space 

```{r}
qf <- qf[,,-c(1:18)]
```

```{r}
#Garbage collection to free space 
gc(); gc()
```

#### Filtering: Remove highly missing precursors

We keep peptides that were observed at last 3 times out of the 
$n= 9$ samples, so that we can estimate the peptide characteristics. 

```{r remove_precursors_with_missingness}
nObs <- 3
n <- ncol(qf[["precursors"]])

(qf <- filterNA(qf, i = c("precursors", "precursorsPTM"), pNA = (n - nObs) / n))
```
```{r}
#Garbage collection to free space 
gc(); gc()
```

#### Filter one-hit wonders

We normally remove proteins that can only be found by one peptide, as such proteins 
may not be trustworthy. 

Here, we opt not to do this because we prefer a protein-level summary based on 
on precursor so that we can estimate usages later on. 

We leave the code here for your reference. 
But commented it out. 

1. We first calculate how many distinct peptides map to each protein group (`PG_ProteinGroups`). We use the stripped precursor sequence, i.e. sequence of 
the base peptide for this purpose. 

2. We store this information in the row data of the precursors assay

3. We filter precursors of one-hit wonder proteins.  


```{r filter_one_hit_wonders}
## Filter for peptides per protein
# pepsPerProtDf <- qf[["precursors"]] |> 
#    rowData() |> 
#    data.frame() |>
#    dplyr::select("Stripped.Sequence", "Protein.Group") |>
#    group_by(Protein.Group) |>
#    mutate(pepsPerProt = Stripped.Sequence |>
#               unique() |> length()
#           ) #1.

# rowData(qf[["precursors"]])$pepsPerProt <- pepsPerProtDf$pepsPerProt #2. 

#qf <- filterFeatures(qf, 
#                              ~ pepsPerProt > 1, 
#                              keep = TRUE) #3.
```


```{r}
#Garbage collection to free space 
gc(); gc()
```

### Log-transformation 


We perform log2-transformation with `logTransform()` from the
`QFeatures` package. We use `base = 2` and store the result in a new 
summarized experiment named `precursors_log`.

```{r log_transformation}
qf <- logTransform(qf, 
                            base = 2, 
                            i = "precursors", 
                            name = "precursors_log")
qf <- logTransform(qf, 
                            base = 2, 
                            i = "precursorsPTM", 
                            name = "precursorsPTM_log")
```

### Normalisation {#sec-norm}

We first evaluate the distributions of the non-normalised precursors per sample.

```{r}
qf[, , "precursors_log"] |> #1.
  longForm(colvars = colnames(colData(qf))) |>  #2. 
  data.frame() |> 
  filter(!is.na(value)) |>
  ggplot() + #3.
  aes(x = value,
      colour = strain,
      group = colname) +
  geom_density() +
  labs(subtitle = "precursors") +
  theme_minimal()
  
qf[, , "precursorsPTM_log"] |> #1.
  longForm(colvars = colnames(colData(qf))) |>  #2. 
  data.frame() |> 
  filter(!is.na(value)) |>
  ggplot() + #3.
  aes(x = value,
      colour = strain,
      group = colname) +
  geom_density() +
  labs(subtitle = "PTM - precursors") +
  theme_minimal()
```


The data seems very clean. However, normalisation is still required. 

We use the median of ratio normalisation, which addresses both 
differences in loading as well as difference in compositionality.

- This first calculates a reference sample by using the log2 geometric mean, i.e. 
the arithmetic mean on a log scale, for every feature.
- Subsequently log2 fold changes are calculated between each feature of a sample 
and the reference sample.
- Finally, the median of these log2 fold changes are used as the log2-tranformed 
normalisation factor. 

The method is the default method for normalisation in bulk RNA-seq data analysis 
with DESeq2. In our hands it performs very well in a proteomics context.

```{r normalisation}
qf <- sweep( #4. Subtract log2 norm factor column-by-column (MARGIN = 2)
  qf,
  MARGIN = 2,
  STATS = nfLogMedianOfRatios(qf,"precursors_log"),
  i = "precursors_log",
  name = "precursors_norm"
)
qf <- sweep( #4. Subtract log2 norm factor column-by-column (MARGIN = 2)
  qf,
  MARGIN = 2,
  STATS = nfLogMedianOfRatios(qf,"precursorsPTM_log"),
  i = "precursorsPTM_log",
  name = "precursorsPTM_norm"
)
```


We explore the effect of the global normalisation in the subsequent plot. 

Formally, the function applies the following operation on each sample
$i$ across all precursors $p$:

$$
y_{ip}^{\text{norm}} = y_{ip} - \log_2(nf_i)
$$

with $y_ip$ the log2-transformed intensities and $nf_i$ the log2-transformed norm factor. 

Upon normalisation, we can see that the distribution of the 
$y_{ip}^{\text{norm}}$ nicely overlap (using the same code as above)

```{r assess_normalisation_all}
qf[, , "precursors_norm"] |>
  longForm(colvars = colnames(colData(qf))) |> 
  data.frame() |> 
  filter(!is.na(value)) |>
  ggplot() + 
  aes(x = value,
      colour = strain,
      group = colname) +
  geom_density() +
  labs(subtitle = "Median normalisation precursors") +
  theme_minimal()
qf[, , "precursorsPTM_norm"] |>
  longForm(colvars = colnames(colData(qf))) |> 
  data.frame() |> 
  filter(!is.na(value)) |>
  ggplot() + 
  aes(x = value,
      colour = strain,
      group = colname) +
  geom_density() +
  labs(subtitle = "Median normalisation PTM precursors") +
  theme_minimal()
```

We also explore the normalisation for the common precursors. 


```{r assess_normalisation_common}
precursorsCommon <- assay(qf[, , "precursors_log"]) |> na.exclude() |> rownames()
qf[precursorsCommon, , "precursors_norm"] |>
  longForm(colvars = colnames(colData(qf))) |> 
  data.frame() |> 
  filter(!is.na(value)) |>
  ggplot() + 
  aes(x = value,
      colour = strain,
      group = colname) +
  geom_density() +
  theme_minimal() +
   labs(subtitle = "Median normalised log2 precursor intensities (common)")

precursorsCommonPTM <- assay(qf[, , "precursorsPTM_log"]) |> na.exclude() |> rownames()
qf[precursorsCommonPTM, , "precursorsPTM_norm"] |>
  longForm(colvars = colnames(colData(qf))) |> 
  data.frame() |> 
  filter(!is.na(value)) |>
  ggplot() + 
  aes(x = value,
      colour = strain,
      group = colname) +
  geom_density() +
  theme_minimal() +
   labs(subtitle = "Median normalised log2 precursor intensities (common)")
```

Remove temporarily objects 

```{r}
rm(precursorsCommon, precursorsCommonPTM)
```

```{r}
#Garbage collection to free space 
gc(); gc()
```

Note, that normalisation could have been avoided by using Precursor.Normalised 
which is already internally normalised by DIA-NN. However, we do not know the 
exact procedure for this. 

This generally works well for many datasets. 

In this vignette we have chosen to conduct the normalisation explicitly ourselves
using a method for which we know the underlying rationale. 

### Summarisation

Here, we summarise the precursor-level data for the non-enriched runs into 
protein intensities using maxLFQ, which is one of the default methods used in 
DIA data analysis.

Note, that the maxLFQ implementation can become slow for large datasets. In that 
case we recommend the median polish method.

- maxLFQ first calculates all pairwise log ratio's between samples only using 
their shared precursors.
  Particularly, it uses the median of the log ratio's between the shared 
  precursors, i.e. 
  $$r_{ij} = median(y_{sj}-y_{si})$$  
  Hence, it first eliminates ion effects. 

- It then estimates the summaries by solving 

$$
\sum_i\sum_j(y^{prot}_j - y^{prot}_i - r_ij)^2
$$

It is implemented in the `maxLFQ` function of the `iq` package. 

`aggregateFeatures()` in the QFeatures package streamlines summarisation. 
It requires the name of a `rowData` column to group the precursors into proteins 
(or protein groups), here `Protein.Group`. We provide the summarisation
approach through the `fun` argument. 

The function will return a `QFeatures` object with a new set that we called 
`proteins`.

Other summarisation methods are available from the `MsCoreUtils` package, see `?aggregateFeatures` for a comprehensive list. 



```{r summarisation, warning=FALSE}
(qf <- aggregateFeatures(
  qf, i = "precursors_norm", 
  name = "proteins",
  fcol = "Protein.Group", 
  fun = function(X) iq::maxLFQ(X)$estimate
))
```

```{r}
#Garbage collection to free space 
gc(); gc()
```

## Data exploration and QC 

Data exploration aims to highlight the main sources of variation in
the data prior to data modelling and can pinpoint to outlying or
off-behaving samples. 

We first performance the QC for the Phospho-enriched runs then for the non-enriched
runs. 

### Phospho-enriched runs

#### Marginal distribution at precursor level 


```{r qc_ptm_density}
qf[, , "precursorsPTM_norm"] |>
  longForm(colvars = colnames(colData(qf))) |> 
  data.frame() |> 
  filter(!is.na(value)) |>
  ggplot() + 
  aes(x = value,
      colour = strain,
      group = colname) +
  geom_density() +
  theme_minimal() +
   labs(subtitle = "Normalised log2 PTM precursor intensities")
```

```{r qc_ptm_boxplot}
qf[, , "precursorsPTM_norm"] |>
  longForm(colvars = colnames(colData(qf))) |> 
  data.frame() |> 
  filter(!is.na(value)) |>
  ggplot() + 
  aes(x = sampleId,
      y = value,
      colour = strain,
      group = colname) +
  xlab("sample") +
  geom_boxplot() +
  theme_minimal() +
   labs(subtitle = "Normalised log2 PTM precursor intensities")
```

```{r}
#Garbage collection to free space 
gc(); gc()
```

#### Charge state 

```{r qc_ptm_chargestate}
qf[, , "precursorsPTM_norm"] |>
  longForm(colvars = colnames(colData(qf)), rowvars = "Precursor.Charge") |>
  as.data.frame() |>
  filter(!is.na(value)) |>
  filter(Precursor.Charge<=4) |> 
ggplot(aes(x = sampleId)) +
  geom_bar(aes(fill = factor(Precursor.Charge, levels = 4:1)),
           colour = "black") +
  labs(title = "Number of phospho-ions per sample",
       x = "Sample",
       fill = "Charge state") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90))
```

```{r}
#Garbage collection to free space 
gc(); gc()
```

#### Identifications per sample 

```{r qc_ptm_id}
qf[,,"precursorsPTM_norm"] |> 
  longForm(colvars = colnames(colData(qf)), 
           rowvars= c("Precursor.Id", 
                      "Protein.Group")) |>
  data.frame() |>
  filter(!is.na(value)) |>
  group_by(strain, sampleId) |>
  summarise(Precursors = length(unique(Precursor.Id)),
            `Protein Groups` = length(unique(Protein.Group))) |> 
  pivot_longer(-(1:2),
               names_to = "Feature",
               values_to = "IDs") |> 
  ggplot(aes(x = sampleId, y = IDs, fill = strain)) +
  geom_col() +
  #scale_fill_observable() +
  facet_wrap(~Feature,
             scales = "free_y") +
  labs(title = "Precursor and protein group identificiations per PTM sample",
       x = "Sample",
       y = "Identifications") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90))
# Data Modeling (Robust Regression){#sec-modelling}
```

```{r}
#Garbage collection to free space 
gc(); gc()
```

#### Dimensionality reduction plot 

A common approach for data exploration is to
perform dimension reduction, such as Multi Dimensional Scaling (MDS).
We will first extract the set to explore along the sample annotations
(used for plot colouring).

```{r qc_ptm_mds}
getWithColData(qf, "precursorsPTM_norm") |> 
  as("SingleCellExperiment") |> 
  runMDS(exprs_values = 1) |> 
  plotMDS(colour_by = "strain") 
```

```{r}
#Garbage collection to free space 
gc(); gc()
```

### Non-enriched runs

#### Marginal distribution at precursor and protein level 

```{r qc_ne_density}
qf[, , "precursors_norm"] |>
  longForm(colvars = colnames(colData(qf))) |> 
  data.frame() |> 
  filter(!is.na(value)) |>
  ggplot() + 
  aes(x = value,
      colour = strain,
      group = colname) +
  geom_density() +
  theme_minimal() +
   labs(subtitle = "Normalised log2 precursor intensities")
```

```{r qc_ne_boxplot}
qf[, , "precursors_norm"] |>
  longForm(colvars = colnames(colData(qf))) |> 
  data.frame() |> 
  filter(!is.na(value)) |>
  ggplot() + 
  aes(x = sampleId,
      y = value,
      colour = strain,
      group = colname) +
  xlab("sample") +
  geom_boxplot() +
  theme_minimal() +
   labs(subtitle = "Normalised log2 precursor intensities")
```

```{r qc_ne_prot_density}
qf[, , "proteins"] |>
  longForm(colvars = colnames(colData(qf))) |> 
  data.frame() |> 
  filter(!is.na(value)) |>
  ggplot() + 
  aes(x = value,
      colour = strain,
      group = colname) +
  geom_density() +
  theme_minimal() +
   labs(subtitle = "log2 protein intensities (normalised precursors)")
```

```{r qc_ne_prot_boxplot}
qf[, , "proteins"] |>
  longForm(colvars = colnames(colData(qf))) |> 
  data.frame() |> 
  filter(!is.na(value)) |>
  ggplot() + 
  aes(x = sampleId,
      y = value,
      colour = strain,
      group = colname) +
  xlab("sample") +
  geom_boxplot() +
  theme_minimal() +
   labs(subtitle = "log2 protein intensities (normalised precursors)")
```

```{r}
#Garbage collection to free space 
gc(); gc()
```

#### Charge state 

```{r qc_ne_chargestate}
qf[, , "precursors_norm"] |>
  longForm(colvars = colnames(colData(qf)), rowvars = "Precursor.Charge") |>
  as.data.frame() |>
  filter(!is.na(value)) |>
  filter(Precursor.Charge<=4) |> 
ggplot(aes(x = sampleId)) +
  geom_bar(aes(fill = factor(Precursor.Charge, levels = 4:1)),
           colour = "black") +
  labs(title = "Number of precursor ions per sample",
       x = "Sample",
       fill = "Charge state") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90))
```

```{r}
#Garbage collection to free space 
gc(); gc()
```

#### Identifications per sample 

```{r qc_ne_id}
qf[,,"precursors_norm"] |> 
  longForm(colvars = colnames(colData(qf)), 
           rowvars= c("Precursor.Id", 
                      "Protein.Group")) |>
  data.frame() |>
  filter(!is.na(value)) |>
  group_by(strain, sampleId) |>
  summarise(Precursors = length(unique(Precursor.Id)),
            `Protein Groups` = length(unique(Protein.Group))) |> 
  pivot_longer(-(1:2),
               names_to = "Feature",
               values_to = "IDs") |> 
  ggplot(aes(x = sampleId, y = IDs, fill = strain)) +
  geom_col() +
  #scale_fill_observable() +
  facet_wrap(~Feature,
             scales = "free_y") +
  labs(title = "Precursor and protein group identificiations per sample",
       x = "Sample",
       y = "Identifications") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90))

```

```{r}
#Garbage collection to free space 
gc(); gc()
```

#### Dimensionality reduction plot 


```{r qc_ne_prot_mds}
getWithColData(qf, "proteins") |> 
  as("SingleCellExperiment") |> 
  runMDS(exprs_values = 1) |> 
  plotMDS(colour_by = "strain") 
```



```{r}
#Garbage collection to free space 
gc(); gc()
```



## Data Modeling at phospho-precursor-level 

### Model estimation

1. We first define the model. We only have one source of variability in the 
experiment that we can model, i.e. the effect of the strain. 

2. We fit the model to each phosphorylated precursor in the `precursorsPTM_norm` 
summarised experiment of the QFeatures object `qf` using the msqrob function. 

```{r model, warning=FALSE}
model <- ~ strain
qf <- msqrob(
  qf,
  i = "precursorsPTM_norm",
  formula = model,
  robust = TRUE)
```

We enabled M-estimation (`robust = TRUE`) for improved robustness
against outliers. 

```{r}
#Garbage collection to free space 
gc(); gc()
```

### Inference

We can now convert the research question "Is the abundance of phospho-precursors 
different between the strains" into a statistical hypotheses. 

In other words, we need to translate this question in a null and alternative 
hypothesis on a single model parameter or a linear combination of model 
parameters, which is also referred to with a contrast.

To aid defining contrasts, we will visualise the experimental design using the `ExploreModelMatrix` package.

```{r design}

vd <- ExploreModelMatrix::VisualizeDesign(
    sampleData =  colData(qf),
    designFormula = model,
    textSizeFitted = 4
)
vd$plotlist
```

We have 3 interesting research questions related to the strain. 

1. Is a precursor DA between ArgSer strain and the WT (control) 
2. Is a precursor DA between ProAla strain and the WT (control)  
3. Is a precursor DA between ProAla strain and the ArgSer strain.  

Each of these contrasts can be translated in a log2 FC or a difference in log2 
FC and can be estimated with one or a linear combination of the model parameters. 

Below we define the contrast and construct the corresponding contrast matrix. 

We can generate the null-hypothesess for all pairwise comparisons manually.
However, here we will use the `createPairwiseContrasts` function in `msqrob2`.

```{r define_H0}
(allHypotheses <- createPairwiseContrasts(
  model, colData(qf), "strain"
  )
)
```

Based on these null hypothesis we now generate the contrast matrix with all
three contrasts using the `makeContrast` function. Note, that experienced users
can also define the contrast matrix, directly. 

```{r make_contrasts}
(L <- makeContrast(
  allHypotheses,
  parameterNames = colnames(vd$designmatrix)
))
```


We assess the contrast for each precursor. 

```{r hypothesis_tests, warning=FALSE}
qf <- hypothesisTest(qf, i = "precursorsPTM_norm", contrast = L, overwrite = TRUE)
```

We extract the results table from the `precursorsPTM_norm` summarised experiment 
in the `qf` object. 

```{r collect_results_PTM}
inferencesPTM <- 
  msqrobCollect(qf[["precursorsPTM_norm"]], L)
```

```{r}
#Garbage collection to free space 
gc(); gc()
```

### Report results

We report the results using a results table, volcano plots and heatmaps. 

#### Results table

We use the 5% nominal FDR level to report results 
```{r}
alpha <- 0.05
```

```{r inference_tables_PTM}
#| results: asis
for (j in colnames(L)) {

  inference <- inferencesPTM |> 
    dplyr::filter(adjPval < alpha & contrast == j)

  cat("**Median - Contrast:**", j, "= 0 (", nrow(inference),
      "significant precursors)\n\n")

  cat('<div style="max-height:300px; overflow-y:auto;">')

  print(
    kable(
      inference |>
        dplyr::arrange(pval) |>
        dplyr::relocate(feature),
      row.names = FALSE
    )
  )

  cat('</div>')

  cat("\n\n\n---\n\n")
}
```

#### Volcanoplots 

```{r}
plotVolcano(inferencesPTM) + 
  facet_wrap(~contrast) + 
  labs(title = "Precursor-level Phospho DA")
```



```{r}
inferencesPTM |> 
  filter(adjPval < alpha) |> 
  mutate(DA = sign(logFC) |> as.factor() |> recode("-1"= "down","1" = "up")) |>
  group_by(contrast, DA) |> 
  ggplot(aes(x = contrast)) +
  geom_bar(aes(fill = factor(DA)),
           colour = "black") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90))
```

#### Heatmaps

We will make a heatmap for each contrast. 
Note that we cluster the rows ourselves to avoid issues with missing values. 

```{r precursor_da_heatmaps}
lapply(colnames(L),
                   function(contrast, se, alpha)
                   {
                     sig <- rowData(se)[[contrast]] |> 
                       filter(adjPval < alpha) |> 
                       rownames()
                     if (length(sig) > 2)
                     {
                     quants <- t(scale(t(assay(se[sig,]))))
                     colnames(quants) <- se$sampleId #specific to this dataset to get short colnames
                     rowclushlp <- quants
                     rowclushlp[is.na(rowclushlp)] <- min(quants,na.rm=TRUE) - 2
                     rowclus <- hclust(dist(rowclushlp))
                     annotations <- columnAnnotation(
                       group = se$strain
                       ) #3.
                     set.seed(1234) ## annotation colours are randomly generated by default
                     return(
                       Heatmap(show_row_names = FALSE,
                       quants, 
                       name = "log2 intensity",
                       top_annotation = annotations, 
                       column_title = paste0(contrast, " = 0"),
                       cluster_rows = rowclus
                       )
                     )
                     } else return(ggplot() + theme_minimal() + ggtitle(paste0(contrast, " = 0")))
                   },
                   se = getWithColData(qf, "precursorsPTM_norm"),
                   alpha = alpha)
```


```{r}
#Garbage collection to free space 
gc(); gc()
```

#### Detail plots

We can explore the data for a precursor to validate the statistical inference 
results. For example, let’s explore the precursor and the normalised 
precursor intensities for the precursor with the most significant log2 fold change.

```{r most_siginificant_PTM}
(target_feature <- inferencesPTM |> 
   dplyr::slice(which.min(pval)) |> 
   pull(feature)
)
inferencesPTM |> 
  filter(feature == target_feature)
```

To obtain the required data, we perform a little data manipulation pipeline:

- We use the QFeatures subsetting functionality to retrieve all data related to 
ion data used for model fitting. 
- We then convert the data with longForm() for plotting. 
- Finally, we plot the log2 normalised intensities for each sample for the 
unnormalised and normalised phospho precursors. 

```{r detail_line_plot}
qf[target_feature, , c("precursorsPTM_log","precursorsPTM_norm")] |> #1
  longForm(colvars = colnames(colData(qf))) |> #2
  data.frame() |> 
  ggplot() +
  aes(x = sampleId,
      y = value) +
  geom_line(aes(group = rowname), linewidth = 0.1) +
  geom_point(aes(colour = strain)) +
  facet_wrap(~ assay, scales = "free") +
  ggtitle(target_feature) +
  theme_minimal() +
  theme(axis.text.x = element_blank())
```

```{r detail_boxplot}
qf[target_feature, , c("precursorsPTM_log","precursorsPTM_norm")] |> #1
  longForm(colvars = colnames(colData(qf))) |> #2
  data.frame() |>
  filter(!is.na(value)) %>% 
  {
  ggplot(.) +
  aes(x = strain,
      y = value) +
  geom_boxplot(aes(colour = strain)) +
  facet_wrap(~ assay, scales = "free") +
  geom_jitter(aes(shape = rowname)) +
  scale_shape_manual(values = seq_len(dplyr::n_distinct(.$rowname))) +
  ggtitle(target_feature) +
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  guides(shape = "none")
  }
```

We prioritised DA phospho-precursors. However, they might be DA due to the 
parent proteins on which the PTMs occur that can also change in abundance 
regardless of the modification. 
Any changes in the abundance of a phospho-peptidoform can thus be confounded 
with changes in protein abundance.

We therefore assess differential abundance at the protein level for the 
non-enriched samples.

```{r}
#Garbage collection to free space 
gc(); gc()
```

## Data Modeling at protein-level (non-enriched samples)

### Model estimation

We can use the same model to model the protein-level data of the non-enriched 
runs. Indeed, they stem from the same biological samples. 

```{r model_ne, warning=FALSE}
qf <- msqrob(
  qf,
  i = "proteins",
  formula = model,
  robust = TRUE)
```

We enabled M-estimation (`robust = TRUE`) for improved robustness
against outliers. 

```{r}
#Garbage collection to free space 
gc(); gc()
```

### Inference

We assess the contrast for each protein. 

```{r inference_ne, warning=FALSE}
qf <- hypothesisTest(qf, i = "proteins", contrast = L, overwrite = TRUE)
```

We extract the results table from the `proteins` summarised experiment in the `qf` object. 

```{r results_ne} 
inferences <- 
  msqrobCollect(qf[["proteins"]], L)
```

```{r}
#Garbage collection to free space 
gc(); gc()
```

### Report results

We report the results using a results table, volcano plots and heatmaps. 

#### Results table


```{r result_tables_ne}
#| results: asis
for (j in colnames(L)) {

  inference <- inferences |> 
    dplyr::filter(adjPval < alpha & contrast == j)

  cat("**Median - Contrast:**", j, "= 0 (", nrow(inference),
      "significant proteins)\n\n")

  cat('<div style="max-height:300px; overflow-y:auto;">')

  print(
    kable(
      inference |>
        dplyr::arrange(pval) |>
        dplyr::relocate(feature),
      row.names = FALSE
    )
  )

  cat('</div>')

  cat("\n\n\n---\n\n")
}
```

#### Volcanoplots 

```{r volcanoplots_ne}
plotVolcano(inferences) + 
  facet_wrap(~contrast) + 
  labs(title = "Non-enriched")
```


```{r sigres_ne}
inferences |> 
  filter(adjPval < alpha) |> 
  mutate(DA = sign(logFC) |> as.factor() |> recode("-1"= "down","1" = "up")) |>
  group_by(contrast, DA) |> 
  ggplot(aes(x = contrast)) +
  geom_bar(aes(fill = factor(DA)),
           colour = "black") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90))
```

We notice that many proteins are DA at the nominal 5% FDR-level. 

#### Heatmaps

```{r precursor_da_heatmaps_ne}
lapply(colnames(L),
                   function(contrast, se, alpha)
                   {
                     sig <- rowData(se)[[contrast]] |> 
                       filter(adjPval < alpha) |> 
                       rownames()
                     quants <- t(scale(t(assay(se[sig,]))))
                     colnames(quants) <- se$sampleId #specific to this dataset to get short colnames
                     rowclushlp <- quants
                     rowclushlp[is.na(rowclushlp)] <- min(quants,na.rm=TRUE) - 2
                     rowclus <- hclust(dist(rowclushlp))
                     annotations <- columnAnnotation(
                       group = se$strain
                       ) #3.
                     set.seed(1234) ## annotation colours are randomly generated by default
                     return(
                       Heatmap(show_row_names = FALSE,
                       quants, 
                       name = "log2 intensity",
                       top_annotation = annotations, 
                       column_title = paste0(contrast, " = 0"),
                       cluster_rows = rowclus
                       )
                     )
                   },
                   se = getWithColData(qf, "proteins"),
                   alpha = alpha)
```

#### Detail plots

We can explore the data for a protein to validate the statistical inference 
results. For example, let’s explore the normalised precursor and the summarised 
protein intensities for the protein with the most significant log2 fold change.

```{r}
(target_feature <- inferences |> 
   dplyr::slice(which.min(pval)) |> 
   pull(feature)
)
inferences |> 
  filter(feature == target_feature)
```

To obtain the required data, we perform a little data manipulation pipeline:

- We use the QFeatures subsetting functionality to retrieve all data related to 
and focusing on the precursors_log and proteins sets that contains the peptide 
ion data used for model fitting. 
- We then convert the data with longForm() for plotting. 
- Finally, we plot the log2 normalised intensities for each sample at the 
protein and at the peptide level. Since multiple peptides are recorded for the 
protein, we link peptides across samples using a grey line. Samples are colored 
according to strain.

```{r}
qf[target_feature, , c("precursors_log","precursors_norm", "proteins")] |> #1
  longForm(colvars = colnames(colData(qf))) |> #2
  data.frame() |> 
  ggplot() +
  aes(x = colname,
      y = value) +
  geom_line(aes(group = rowname), linewidth = 0.1) +
  geom_point(aes(colour = strain)) +
  facet_wrap(~ assay, scales = "free") +
  ggtitle(target_feature) +
  theme_minimal() +
  theme(axis.text.x = element_blank())
```

```{r warning = FALSE}
qf[target_feature, , c("precursors_log","precursors_norm", "proteins")] |> #1
  longForm(colvars = colnames(colData(qf))) |> #2
  data.frame() |>
  filter(!is.na(value)) %>% 
  {
  ggplot(.) +
  aes(x = strain,
      y = value) +
  geom_boxplot(aes(colour = strain)) +
  facet_wrap(~ assay, scales = "free") +
  geom_jitter(aes(shape = rowname)) +
  scale_shape_manual(values = seq_len(dplyr::n_distinct(.$rowname))) +
  ggtitle(target_feature) +
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  guides(shape = "none")
  }
```



#### Compare phospo-precursor level intensities to protein level intensities
 
We compare the intensities for the top 5 DA phospo-precursors for the first 
contrast to these of their corresponding protein in the non-enriched assay. 
 
```{r comparison_plots, warning = FALSE, message = FALSE}
contr = colnames(L)[1]
top5 <- inferencesPTM |> 
  filter(contrast == contr) |> 
  arrange(pval) |> 
  head(n = 5) |> 
  pull(feature)

for (feat in top5)
{
  ptm_data <- qf[,,c("precursorsPTM_norm")] |> 
    longForm(colvars = colnames(colData(qf)), rowvars = "Protein.Group") |>
    data.frame() |> 
    filter(rowname==feat)
  feature_protein <- ptm_data |> 
    pull("Protein.Group") |> 
    unique()
  prot_data <- qf[,,"proteins"]|>
    longForm(colvars = colnames(colData(qf)), rowvars = "Protein.Group") |>
    data.frame() |> 
    filter(Protein.Group==feature_protein) 
  ptm_protein <- rbind(ptm_data, prot_data)
  ylims <- ptm_protein |> 
  group_by(assay) |> 
  summarise(cent = mean(range(value,na.rm=TRUE)), ampl = diff(range(value,na.rm=TRUE))) |> 
  mutate(lower = cent - max(ampl)/2, 
         upper = cent + max(ampl)/2) |> 
  select(-c(cent, ampl))
  
  comparison_plot <- ptm_protein |>
    ggplot() +
    aes(x = sampleId,
        y = value) +
    geom_line(aes(group = rowname), linewidth = 0.1) +
    geom_point(aes(colour = strain)) +
    facet_wrap(~ assay, scales = "free") +
    labs(
      title = paste0(feat," / ", feature_protein),
      subtitle = paste("padj PTM =", 
                       inferencesPTM |> 
                         filter(feature == feat & contrast == contr) |> 
                         pull(adjPval) |> 
                         round(digits = 3))
    ) +
    theme_minimal() +
    theme(axis.text.x = element_blank()) +
    ggh4x::facetted_pos_scales(
    y = list(
      assay == ylims$assay[1] ~ scale_y_continuous(limits = unlist(ylims[1,c("lower","upper")])),
      assay == ylims$assay[2] ~ scale_y_continuous(limits = unlist(ylims[2,c("lower","upper")])),
      assay == ylims$assay[3] ~ scale_y_continuous(limits = unlist(ylims[3,c("lower","upper")]))
    )
    ) 
 print(comparison_plot)
}
```

```{r}
#Garbage collection to free space 
gc(); gc()
```

## Data Modeling of phospho-peptidoform usages

### Background 

- Differential abundant PTMs can be due protein abundance

- Port concept of differential usage towards PTMs: 

$$\log_2 DU = \log_2 FC^\text{PTM} - log_2 FC^\text{protein}$$

Two designs: PTM and protein level quantification be done 

- on same sample: paired design 
- using independent samples


#### Paired samples 

1. Perform regular data processing at the protein-level: 
  
    - log2-transformation, 
    - normalisation and 
    - summarisation

2. Calculate relative log2 transformed abundances at peptidoform level 
$$y^\text{pep,rel}_i = y^\text{pep}_i - y^\text{prot}_i$$

3. Perform conventional differential msqrob2 analysis on the relative 
peptidoform level. $\rightarrow$ Immediately gives us diferential usages!
    
4. Optionally 
    
  a. aggregate relative log2 transformed abundances at peptidoform level at 
     PTM-level
  b. perform a differential msqrob2 analysis at PTM-level


#### Unpaired PTM-level and protein samples 

This is the default MSstatsTMT approach, which they also use for paired designs?!

  1. Perform regular msqrob2 differential analysis at the protein-level
  2. Perform regular msqrob2 differential analysis at the peptidoform (PTM-level)
  3. Calculate usages
  
$$ 
\begin{array}{ccc}
\log_2 \widehat{DU} & =& \log_2 \widehat{FC}^\text{PTM} - log_2 \widehat{FC}^\text{protein}\\
SE_\widehat{DU} &=& \sqrt{(SE_\widehat{FC}^{PTM})^2 + (SE_\widehat{FC}^{protein})^2}
\end{array}
$$

#### Our study

For this dataset we have phospho-enriched and non-enriched runs on the same 
biological samples. 

We subsetted the data to have three unpaired samples for the phospho-enriched 
and non-enriched runs for each straing. So we do the Unpaired analysis.


### Unpaired samples: calculate usages

We first define a function to correct phospho-precursor-level log-FC with 
non-enriched logFC.

1. Matches the phospho-precursor to the non-enriched Protein.Group
2. For each contrast

- 2.1. Extracts the inference results for the contrast at tc protein level
- 2.2. Adds them to the inference results for the contrast at the lip precursor level
- 2.3. Calculates the lip relative differential abundance (RDA), i.e. the logFC corrected for the tc logFC at the protein level and the corresponding inference

```{r}
calculate_usage <- function(qf, 
                          i_ptm, 
                          i_ne, 
                          contrasts, 
                          fcol_ptm = "Protein.Group", 
                          fcol_ne = "Protein.Group",  
                          satterthwaite = FALSE)
{
  match_precursor_to_protein_FC <- match(rowData(qf[[i_ptm]])[[fcol_ptm]],
                                         rowData(qf[[i_ne]])[[fcol_ne]]) #1. 
  for (contrast in contrasts)
    {
    results_ne <- rowData(qf[[i_ne]])[[contrast]] #2.1
    colnames(results_ne) <- results_ne |> 
      colnames() |> 
      paste0("_ne")
    rowData(qf[[i_ptm]])[[contrast]] <- cbind(
      rowData(qf[[i_ptm]])[[contrast]], 
      results_ne[match_precursor_to_protein_FC,]) |> #2.2 
      mutate(
        logUsage = logFC - logFC_ne, #2.3
        se_usage = sqrt(se^2 + se_ne^2),
        t_usage = logUsage /se_usage,
        df_usage = if (satterthwaite) (se^2 + se_ne^2)^2/(se^4/df + se_ne^4/df_ne) else df,
        pval_usage = pt(abs(t_usage), df_usage, lower.tail = FALSE)*2, 
        adjPval_usage = p.adjust(pval_usage, method = "fdr")
        )
  }
    return(qf)
}
```

Calculate log-transformed usages (relative differential abundances).

```{r}
qf <- calculate_usage(qf, i_ptm = "precursorsPTM_norm", i_ne = "proteins", contrasts = colnames(L))
```

We extract the results table from the `precursorsPTM_norm` summarised experiment 
in the `qf` object. 

```{r collect_results_PTM_usage}
inferencesUsage <- 
  msqrobCollect(qf[["precursorsPTM_norm"]], L) |> 
  relocate(contains("sage"))
```

### Report results

We report the results using a results table, volcano plots and heatmaps. 

#### Results table

```{r result_tables_usage}
#| results: asis
for (j in colnames(L)) {

  inference <- inferencesUsage |> 
    dplyr::filter(adjPval_usage < alpha & contrast == j)

  cat("**Median - Contrast:**", j, "= 0 (", nrow(inference),
      "significant proteins)\n\n")

  cat('<div style="max-height:300px; overflow-y:auto;">')

  print(
    kable(
      inference |>
        dplyr::arrange(pval) |>
        dplyr::relocate(feature),
      row.names = FALSE
    )
  )

  cat('</div>')

  cat("\n\n\n---\n\n")
}
```

#### Volcanoplots 

```{r}
inferencesUsage |> 
  filter(is.finite(adjPval_usage)) |> 
  ggplot(aes(x = logUsage, 
             y = -log10(pval_usage), 
             col = adjPval_usage < alpha)) +
    geom_point(size = 1, alpha = 0.5) + 
    theme_bw() + 
    ylab("-log10(p-value)") + 
    scale_color_manual(values = c(`FALSE` = "black", `TRUE` = "red"), 
                       breaks = c("TRUE", "FALSE")) + 
    labs(color = paste0("FDR < ", alpha)) + 
    facet_wrap(~contrast) + 
    labs(title = "Usage")
```


```{r}
inferencesUsage |> 
  filter(adjPval_usage < alpha) |> 
  mutate(DA = sign(logUsage) |> as.factor() |> recode("-1"= "down","1" = "up")) |>
  group_by(contrast, DA) |> 
  ggplot(aes(x = contrast)) +
  geom_bar(aes(fill = factor(DA)),
           colour = "black") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90))
```

We observe that the number of significant precursors reduced. 


#### Detail plots

We can explore the data for a protein to validate the statistical inference 
results. For example, let’s explore the normalised precursor abundance
as well as the summarised protein intensities for the protein with the most 
significant log2 fold change.

```{r}
(target_feature <- inferencesUsage |> 
   dplyr::slice(which.min(pval_usage)) |> 
   pull(feature)
)
inferencesUsage |> 
  filter(feature == target_feature)
```

We again make a similar plot as before.

```{r}
ptm <-  qf[target_feature, , c("precursorsPTM_norm")] |> #1
  longForm(colvars = colnames(colData(qf)), rowvars = "Protein.Group") |> #2
  data.frame()
protein <- qf[unique(ptm$Protein.Group), , c("proteins")] |> #1
  longForm(colvars = colnames(colData(qf)), rowvars = "Protein.Group") |> #2
  data.frame()


ptm_protein <- rbind(ptm, protein)
ylims <- ptm_protein |> 
  group_by(assay) |> 
  summarise(cent = mean(range(value,na.rm=TRUE)), ampl = diff(range(value,na.rm=TRUE))) |> 
  mutate(lower = cent - max(ampl)/2, 
         upper = cent + max(ampl)/2) |> 
  select(-c(cent, ampl))

outplot <- ptm_protein |>
  ggplot() +
  aes(x = colname,
      y = value) +
  geom_line(aes(group = rowname), linewidth = 0.1) +
  geom_point(aes(colour = strain)) +
  facet_wrap(~ assay, scales = "free") +
  ggh4x::facetted_pos_scales(
    y = list(
      assay == ylims$assay[1] ~ scale_y_continuous(limits = unlist(ylims[1,c("lower","upper")])),
      assay == ylims$assay[2] ~ scale_y_continuous(limits = unlist(ylims[2,c("lower","upper")]))
    )
    ) +
  labs(subtitle = target_feature) +
  theme_minimal() +
  theme(axis.text.x = element_blank())
print(outplot)
```

```{r}
rbind(ptm, protein) %>%
  {
  ggplot(.) +
  aes(x = strain,
      y = value) +
  geom_boxplot(aes(colour = strain)) +
  facet_wrap(~ assay, scales = "free") +
  geom_jitter(aes(shape = rowname)) +
  scale_shape_manual(values = seq_len(dplyr::n_distinct(.$rowname))) +
  ggtitle(target_feature) +
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  guides(shape = "none")
  }
```

We compare the intensities for the top 5 DA phospo-precursors for the first 
contrast to these of their corresponding protein in the non-enriched assay and
to the usages. 
 
```{r comparison_plots_top5, warning = FALSE, message = FALSE}
contr = colnames(L)[1]
top5 <- inferencesPTM |> 
  filter(contrast == contr) |> 
  arrange(pval) |> 
  head(n = 5) |> 
  pull(feature)

for (feat in top5)
{
  ptm_data <- qf[,,c("precursorsPTM_norm")] |> 
    longForm(colvars = colnames(colData(qf)), rowvars = "Protein.Group") |>
    data.frame() |> 
    filter(rowname==feat)
  feature_protein <- ptm_data |> 
    pull("Protein.Group") |> 
    unique()
  prot_data <- qf[,,"proteins"]|>
    longForm(colvars = colnames(colData(qf)), rowvars = "Protein.Group") |>
    data.frame() |> 
    filter(Protein.Group==feature_protein) 
  ptm_protein <- rbind(ptm_data, prot_data)
  ylims <- ptm_protein |> 
  group_by(assay) |> 
  summarise(cent = mean(range(value,na.rm=TRUE)), ampl = diff(range(value,na.rm=TRUE))) |> 
  mutate(lower = cent - max(ampl)/2, 
         upper = cent + max(ampl)/2) |> 
  select(-c(cent, ampl))
  
  comparison_plot <- ptm_protein |>
    ggplot() +
    aes(x = sampleId,
        y = value) +
    geom_line(aes(group = rowname), linewidth = 0.1) +
    geom_point(aes(colour = strain)) +
    facet_wrap(~ assay, scales = "free") +
    labs(
      title = paste0(feat," / ", feature_protein),
      subtitle = paste("padj PTM =", 
                       inferencesPTM |> 
                         filter(feature == feat & contrast == contr) |> 
                         pull(adjPval) |> 
                         round(digits = 3),
                       "padj usage =", 
                       inferencesUsage |> 
                         filter(feature == feat & contrast == contr) |> 
                         pull(adjPval_usage) |>
                         round(digits = 3))
    ) +
    theme_minimal() +
    theme(axis.text.x = element_blank()) +
    ggh4x::facetted_pos_scales(
    y = list(
      assay == ylims$assay[1] ~ scale_y_continuous(limits = unlist(ylims[1,c("lower","upper")])),
      assay == ylims$assay[2] ~ scale_y_continuous(limits = unlist(ylims[2,c("lower","upper")]))
    )
    ) 
 print(comparison_plot)
}
```

We compare the intensities for the top 5 DU phospo-precursors for the first 
contrast to these of their corresponding protein in the non-enriched assay and
to their unadjusted abundances. 
 
```{r comparison_plots_top5_DU, warning = FALSE, message = FALSE}
contr = colnames(L)[1]
top5 <- inferencesUsage |> 
  filter(contrast == contr) |> 
  arrange(pval_usage) |> 
  head(n = 5) |> 
  pull(feature)

for (feat in top5)
{
  ptm_data <- qf[,,c("precursorsPTM_norm")] |> 
    longForm(colvars = colnames(colData(qf)), rowvars = "Protein.Group") |>
    data.frame() |> 
    filter(rowname==feat)
  feature_protein <- ptm_data |> 
    pull("Protein.Group") |> 
    unique()
  prot_data <- qf[,,"proteins"]|>
    longForm(colvars = colnames(colData(qf)), rowvars = "Protein.Group") |>
    data.frame() |> 
    filter(Protein.Group==feature_protein) 
  ptm_protein <- rbind(ptm_data, prot_data)
  ylims <- ptm_protein |> 
  group_by(assay) |> 
  summarise(cent = mean(range(value,na.rm=TRUE)), ampl = diff(range(value,na.rm=TRUE))) |> 
  mutate(lower = cent - max(ampl)/2, 
         upper = cent + max(ampl)/2) |> 
  select(-c(cent, ampl))
  
  comparison_plot <- ptm_protein |>
    ggplot() +
    aes(x = sampleId,
        y = value) +
    geom_line(aes(group = rowname), linewidth = 0.1) +
    geom_point(aes(colour = strain)) +
    facet_wrap(~ assay, scales = "free") +
    labs(
      title = paste0(feat," / ", feature_protein),
      subtitle = paste("padj PTM =", 
                       inferencesPTM |> 
                         filter(feature == feat & contrast == contr) |> 
                         pull(adjPval) |> 
                         round(digits = 3),
                       "padj usage =", 
                       inferencesUsage |> 
                         filter(feature == feat & contrast == contr) |> 
                         pull(adjPval_usage) |>
                         round(digits = 3))
    ) +
    theme_minimal() +
    theme(axis.text.x = element_blank()) +
    ggh4x::facetted_pos_scales(
    y = list(
      assay == ylims$assay[1] ~ scale_y_continuous(limits = unlist(ylims[1,c("lower","upper")])),
      assay == ylims$assay[2] ~ scale_y_continuous(limits = unlist(ylims[2,c("lower","upper")])))
    ) 
 print(comparison_plot)
}
```

```{r}
#Garbage collection to free space 
gc(); gc()
```

## Differential usage analysis at the PTM level

Multiple precursors are carrying the same PTM (cover the PTM at a certain 
position in the protein). 
We can aggregate them together. 

1. Here we first aggregate the precursor level abundances
2. Calculate the DA at the PTM-level
3. Correct the DA at PTM-level with the DA at the protein level to obtain the 
   usages

### Data processing 

The variable Protein.Sites in the DIA-NN parquet file contains information on the 
exact modified residue position in the full protein sequence. 

We first will make a new assay with duplicates for the precursors that contain 
multiple phospho sites. 

1. We extract the percursor-level assay 
2. We extract the row data with the original variables (drop msqrob2 colums)
3. Add a new variable with the modification type (21 is phosho)
4. Add a new variable with the location of the modification on the original 
   protein
5. replace the rowData of se
6. Make vector of the row numbers with duplicates for the sequences containing 
   multiple PTMs
7. Make a corresponding vector with the positions of the individual PTM sites
8. Make a corresponding vector with the modification type
9. Extract the ids of rows that correspond to phospho sites
10. Make a new summarised experiment with duplicate entries for precursors with 
    multiple phospho sites. 
11. Add the position of the phospho site
12. Add a variable with the Protein.Group name and the phospho site position 
    attached to it. 
13. Make the rownames unique by appending the phospho site position to it.
14. Add the assay to the QFeatures object.

```{r}
se <- qf[["precursorsPTM_norm"]] #1. 

rd <- rowData(se)[,names(rowData(qf[["precursorsPTM"]]))] |> ##2.
  as.data.frame() |> 
  mutate(
    mods  = str_extract_all(Modified.Sequence, 
                           "UniMod:[0-9]+") |> 
      lapply(FUN = gsub, pattern="UniMod:",replacement=""), ##3.
    pos = str_extract(Protein.Sites, 
                      "(?<=:)[^\\]]+") |> 
      strsplit(split = ",") |> 
      lapply(FUN=substr,start=2,stop=10000) |> 
      lapply(FUN=as.integer) ##4.
  )

rowData(se) <- rd ##5.

ids <- rep(rd |> 
             nrow() |>
             seq_len(), 
           rd |>
             pull(pos) |>
             sapply(FUN = length))##6.
pos <- unlist(rd$pos) ##7.
mod <- unlist(rd$mods) ##8. 

phosIds <- ids[mod == "21"] ##9.


seDup <- se[phosIds,] ##10.
rowData(seDup)$pos <- pos[mod == "21"] ##11.

rowData(seDup)$Protein.Group.Mod <- paste(
  rowData(seDup)$Protein.Group,
  rowData(seDup)$pos, sep = "_") ##12.

rownames(seDup) <- paste(rownames(seDup), rowData(seDup)$pos, sep = "_") ##13.
qf <- addAssay(qf, seDup, "precursorsPTM_norm_unnested") ##14.
```

We now aggregate the data of precursors that map to the same PTM position on a 
protein in one protein expression value. 

```{r}
qf <- aggregateFeatures(qf, 
                   i = "precursorsPTM_norm_unnested",
                   fcol = "Protein.Group.Mod", 
                   name = "ptm",
                   fun = function(X, ...) iq::maxLFQ(X)$estimate
                   )
```

We assess the distribution of the PTM usages. 

```{r}
qf[, , "ptm"] |> #1.
  longForm(colvars = colnames(colData(qf))) |>  #2. 
  data.frame() |> 
  filter(!is.na(value)) |>
  ggplot() + #3.
  aes(x = value,
      colour = strain,
      group = colname) +
  geom_density() +
  labs(subtitle = "PTM") +
  theme_minimal()
```

We remove temporarily objects

```{r}
rm(se, seDup, mod, pos, ids, phosIds)
```

```{r}
#Garbage collection to free space 
gc(); gc()
```

### Model estimation

Again we can estimate the same model but now for the PTM abundances. 

```{r model_ptm_usage, warning=FALSE}
qf <- msqrob(
  qf,
  i = "ptm",
  formula = model,
  robust = TRUE)
```

We enabled M-estimation (`robust = TRUE`) for improved robustness
against outliers. 

```{r}
#Garbage collection to free space 
gc(); gc()
```

### Inference

Contrasts remain the same. 
We assess the contrast for each precursor. 

```{r inference_ptm_usage, warning=FALSE}
qf <- hypothesisTest(qf, i = "ptm", contrast = L)
```

Calculate log-transformed usages (relative differential abundances).

```{r}
qf <- calculate_usage(qf, i_ptm = "ptm", i_ne = "proteins", contrasts = colnames(L))
```

We extract the results table from the `proteins` summarised experiment in the `qf` object. 

```{r results_ptm_usage}
inferencesPtmUsage <- 
  msqrobCollect(qf[["ptm"]], L) |> 
  relocate(contains("sage"))
```

```{r}
#Garbage collection to free space 
gc(); gc()
```

### Report results

We report the results using a results table, volcano plots and heatmaps. 

#### Results table

```{r result_tables_ptm_usage}
#| results: asis
for (j in colnames(L)) {

  inference <- inferencesPtmUsage |> 
    dplyr::filter(adjPval_usage < alpha & contrast == j)

  cat("**Median - Contrast:**", j, "= 0 (", nrow(inference),
      "significant proteins)\n\n")

  cat('<div style="max-height:300px; overflow-y:auto;">')

  print(
    kable(
      inference |>
        dplyr::arrange(pval) |>
        dplyr::relocate(feature),
      row.names = FALSE
    )
  )

  cat('</div>')

  cat("\n\n\n---\n\n")
}
```

#### Volcanoplots 

```{r}
inferencesPtmUsage |> 
  filter(is.finite(adjPval_usage)) |> 
  ggplot(aes(x = logUsage, 
             y = -log10(pval_usage), 
             col = adjPval_usage < alpha)) +
    geom_point(size = 1, alpha = 0.5) + 
    theme_bw() + 
    ylab("-log10(p-value)") + 
    scale_color_manual(values = c(`FALSE` = "black", `TRUE` = "red"), 
                       breaks = c("TRUE", "FALSE")) + 
    labs(color = paste0("FDR < ", alpha)) + 
    facet_wrap(~contrast) + 
    labs(title = "Usage")
```


```{r}
inferencesPtmUsage |> 
  filter(adjPval_usage < alpha) |> 
  mutate(DA = sign(logUsage) |> as.factor() |> recode("-1"= "down","1" = "up")) |>
  group_by(contrast, DA) |> 
  ggplot(aes(x = contrast)) +
  geom_bar(aes(fill = factor(DA)),
           colour = "black") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90))
```


#### Detail plots


We can explore the data for a protein to validate the statistical inference results. For example, let’s explore the precursor, summarized PTM and summarised protein intensities for the PTM with the most significant log2 fold change.

```{r}
(target_feature <- inferencesPtmUsage |> 
   dplyr::slice(which.min(pval_usage)) |> 
   pull(feature)
)
inferencesPtmUsage |> 
  filter(feature == target_feature)
```

```{r}
ptm <-  qf[target_feature, , c("precursorsPTM_norm_unnested", "ptm")] |> #1
  longForm(colvars = colnames(colData(qf)), rowvars = "Protein.Group") |> #2
  data.frame()
protein <- qf[unique(ptm$Protein.Group), , c("proteins")] |> #1
  longForm(colvars = colnames(colData(qf)), rowvars = "Protein.Group") |> #2
  data.frame()


ptm_protein <- rbind(ptm, protein)
ylims <- ptm_protein |> 
  group_by(assay) |> 
  summarise(cent = mean(range(value,na.rm=TRUE)), ampl = diff(range(value,na.rm=TRUE))) |> 
  mutate(lower = cent - max(ampl)/2, 
         upper = cent + max(ampl)/2) |> 
  select(-c(cent, ampl))

ptm_protein |>
  filter(!is.na(value)) |>
  ggplot() +
  aes(x = sampleId,
      y = value) +
  geom_line(aes(group = rowname), linewidth = 0.1) +
  geom_point(aes(colour = strain)) +
  facet_wrap(~ assay, scales = "free") +
  ggh4x::facetted_pos_scales(
    y = list(
      assay == ylims$assay[1] ~ scale_y_continuous(limits = unlist(ylims[1,c("lower","upper")])),
      assay == ylims$assay[2] ~ scale_y_continuous(limits = unlist(ylims[2,c("lower","upper")])),
      assay == ylims$assay[3] ~ scale_y_continuous(limits = unlist(ylims[3,c("lower","upper")]))
    )
    ) +
  labs(subtitle = paste0("Protein:", sub("_"," / phosho-position: ", target_feature))) +
  theme_minimal() +
  theme(axis.text.x = element_blank())
```

```{r}
rbind(ptm, protein) %>%
  {
  ggplot(.) +
  aes(x = strain,
      y = value) +
  geom_boxplot(aes(colour = strain)) +
  facet_wrap(~ assay, scales = "free") +
  geom_jitter(aes(shape = rowname)) +
  scale_shape_manual(values = seq_len(dplyr::n_distinct(.$rowname))) +
  labs(subtitle = paste0("Protein:", sub("_"," / phosho-position: ", target_feature))) +
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  guides(shape = "none")
  }
```

```{r}
#Garbage collection to free space 
gc(); gc()
```

### Along Protein Mapping 

In this section we will make Along Protein graphs. 
Therefore we need sequence information on the protein sequence, which we can 
find in the fasta file from uniprot that was used for mapping. 

Note, that it is recommended to use the fasta that was used for the mapping. 
However, this file was not uploaded by the authors so we use the version of the 
current release for illustration purposes. 

1. Read fasta file (download it from uniprot)

```{r}
fastaFile <- bfcrpath(bfc,"https://rest.uniprot.org/uniprotkb/stream?compressed=true&download=true&format=fasta&query=%28%28proteome%3AUP000002311%29+AND+reviewed%3Dtrue%29")
fasta <- seqinr::read.fasta(fastaFile, seqtype = "AA", as.string = T) #2.
```

2. Extract the proteins from the fasta file.

```{r}
# read in fasta file for protein sequence & length -> to calculate the coverage
mapping <- data.frame(fullName = fasta |> names(), 
                      accession = fasta |> names() |> stringr::str_split_i("\\|", 2),
                      sequence = unlist(fasta)) |> 
  mutate(length = nchar(sequence))
rownames(mapping) <- mapping$accession
```


#### PTM barcode plot 

We make a barplot for all phosphorylated residues on protein 
`r sub("_.*","", target_feature)`

We first gather the data:

1. We extract the protein name 
2. we filter all PTMs for that protein
3. we add the phospho-position in the rowData to it
4. we make a new variable type to indicate if the PTM usage is up, down, 
     non-significant or missing

```{r}
prot <- sub("_.*","", target_feature) ## 1.
ptmListProt <- inferencesPtmUsage |> 
  filter(grepl(prot, feature)) |> ## 2. 
  left_join(rowData(qf[["ptm"]]) |> 
              as.data.frame() |> 
              select(Protein.Group.Mod, pos),
            by = c("feature" = "Protein.Group.Mod")) |> ## 3. 
  mutate(type = ifelse(is.na(adjPval_usage),  
                       "missing", 
                       ifelse(adjPval_usage < 0.05, 
                              ifelse(logUsage < 0, 
                                     "down",
                                     "up"),
                              "non-sig")) ## 4.
  )
```

We make the barplot by adding a geom_point layer and a vertical line (geom_vline) 
to indicate the residue position. 

```{r}
ptmListProt |> 
  ggplot(aes(x=pos,y=0,col=type)) +
  geom_point() +
  geom_vline(aes(xintercept=pos,col=type)) + 
  xlab("residue") + 
  ylab("") +
  scale_color_manual(values = c("black","grey","blue","red"), 
                     breaks = c("missing","non-sig","down", "up")) +
  theme(
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank(), 
    panel.grid = element_blank(),
    panel.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA),
    strip.background = element_rect(color = "black")
    ) +
  xlim(0, mapping[prot,"length"]) +
  ggtitle(paste0(prot, ": Phospho residues")) +
  facet_grid(contrast~.) 
```

#### Along protein plot for all precursors

We first gather the data from all precursors that map to protein 
`r sub("_.*","", target_feature)`

```{r}
prot <- sub("_.*","", target_feature) ## 1.

precursorListProt <- inferencesUsage |> 
  left_join(rowData(qf[["precursorsPTM_norm"]]) |> 
              as.data.frame() |> 
              select(Precursor.Id, Protein.Group, Stripped.Sequence, Protein.Sites, Modified.Sequence),
            by = c("feature" = "Precursor.Id")) |> ## 2. 
  filter(Protein.Group == prot) |> ## 3. 
  mutate(seqOrder = as.factor(feature) |> as.numeric(),
         type = ifelse(is.na(adjPval_usage),  
                       "missing", 
                       ifelse(adjPval_usage < 0.05, 
                              ifelse(logUsage < 0, 
                                     "down",
                                     "up"),
                              "non-sig")), ## 4.
         tmp = str_locate_all(mapping[prot,"sequence"], Stripped.Sequence),
         start = purrr::map(tmp, ~ .x[, "start"]),
         end = purrr::map(tmp, ~ .x[, "end"])) |>
  unnest(c(start, end))
```

Add location of Phospo residues. 

1. Add a new variable with the modification type (21 is phosho)
2. Add a new variable with the location of the modification on the original 
   protein
3. Make vector of the row numbers with duplicates for the sequences containing 
   multiple PTMs
4. Make a corresponding vector with the positions of the individual PTM sites
5. Make a corresponding vector with the modification type
6. Extract the ids of rows that correspond to phospho sites
7. Make a data.frame with duplicate entries for precursors with 
    multiple phospho sites with variables seqOrder, type and contrast 
8. Add the position of the phospho site



```{r}
precursorListProt <-  precursorListProt |> mutate(
    mods  = str_extract_all(Modified.Sequence, 
                           "UniMod:[0-9]+") |> 
      lapply(FUN = gsub, pattern="UniMod:",replacement=""), ##1.
    pos = str_extract(Protein.Sites, 
                      "(?<=:)[^\\]]+") |> 
      strsplit(split = ",") |> 
      lapply(FUN=substr,start=2,stop=10000) |> 
      lapply(FUN=as.integer)) ## 2.

ids <- rep(precursorListProt |> 
             nrow() |>
             seq_len(), 
           precursorListProt |>
             pull(pos) |>
             sapply(FUN = length))## 3.
pos <- unlist(precursorListProt$pos) ## 4.
mod <- unlist(precursorListProt$mods) ## 5. 

phosIds <- ids[mod == "21"] ## 6.


precursorPosListProt <- precursorListProt[phosIds,] |> ## 7.
  select(seqOrder,type, contrast) |>
  mutate(pos = pos[mod == "21"]) ## 8.
```

Make along protein plot. 


```{r}
precursorListProt |> 
  ggplot(aes(y=seqOrder)) +
    geom_segment(aes(x = start, xend = end, yend = seqOrder, color = type),
                  linewidth = 1) + 
    xlim(0, mapping[prot,"length"]) +
  ylab("Precursors") + xlab("Residue") +  #6. 
  scale_y_continuous(breaks = precursorListProt$seqOrder, labels = precursorListProt$Stripped.Sequence) +
  scale_color_manual(values = c("black","grey","blue","red", "black"), 
                     breaks = c("missing","non-sig","down", "up", "black")) +
  geom_segment(data = precursorPosListProt,
               aes(x = pos, xend = pos, y = seqOrder - 0.5, yend = seqOrder + 0.5, 
                   color = "black"), linewidth = 1) +
  facet_grid(contrast~.) +
  theme(
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank(), #We suppress the feature sequence here because there are too many sequences
    panel.grid = element_blank(),
    panel.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA),
    strip.background = element_rect(color = "black")
    ) +
    labs(title = paste0(prot, " along with precursors"),
           subtitle = "(phospho residue in black vertical line)") 

```



