# Useful links and session information{#sec-end}

```{r, echo = FALSE}
source("R/knitr_setup.R")
```

## Data sets

We refer here the data sources used in the book:

### E. Coli LFQ spike-in data set

Original study: Shen X, Shen S, Li J, Hu Q, Nie L, Tu C, et al. (2018)
Ionstar enables high-precision, low-missing-data proteomics
quantification in large bio- logical cohorts. Proc.  Natl.  Acad.
Sci. U.S.A. 115, E4767–E4776

Reanalysis study: Sticker A, Goeminne L, Martens L, Clement L. Robust
Summarization and Inference in Proteome-wide Label-free
Quantification. Mol Cell Proteomics. 2020;19(7):1209-1219.

Link to data: https://github.com/statOmics/MSqRobSumPaper/raw/refs/heads/master/spikein/data/maxquant/peptides.zip

Link to data in archive: **TODO** add link to Zenodo

Used in chapters \@ref(sec-basics), and \@ref(sec-benchmarking).

### TMT spike-in data set

Original study: Huang T, Choi M, Tzouros M, Golling S, Pandya NJ,
Banfai B, et al. MSstatsTMT: Statistical Detection of Differentially
Abundant Proteins in Experiments with Isobaric Labeling and Multiple
Mixtures. Mol Cell Proteomics. 2020;19(10):1706-1723.

Reanalysis study: Vandenbulcke S, Vanderaa C, Crook O, Martens L,
Clement L. Msqrob2TMT: Robust linear mixed models for inferring
differential abundant proteins in labeled experiments with arbitrarily
complex design. Mol Cell Proteomics. 2025;24(7):101002.
  
Data source: MassIVE repository (RMSV000000265)

Link to data from archive: https://zenodo.org/records/14767905

Used in chapter \@ref(sec-advanced).

### Francisella data set



### Heart data set


### Mouse diet data set


## License

This book is distributed under a
[Artistic-2.0](https://opensource.org/license/artistic-2-0) license.

**TODO**: add logo and explicit how people can access and modify this work

## Citation

Please cite this book as:

**TODO**: add citation once published

Please cite the `msqrob2` package as: 

> Goeminne L, Gevaert K, Clement L (2016). "Peptide-level Robust 
  Ridge Regression Improves Estimation, Sensitivity, and Specificity
  in Data-dependent Quantitative Label-free Shotgun Proteomics."
  _Molecular & Cellular Proteomics_, *15*(2), 657-668.
  doi:[10.1074/mcp.m115.055897](https://doi.org/10.1074/mcp.m115.055897).

If you opt for a summarised based workflow, you can also cite: 

> Sticker A, Goeminne L, Martens L, Clement L (2020). "Robust 
  Summarization and Inference in Proteome-wide Label-free
  Quantification." _Molecular & Cellular Proteomics_, *19*(7),
  1209-1219.
  doi:[10.1074/mcp.ra119.001624](https://doi.org/10.1074/mcp.ra119.001624).

If you use TMT-based workflows, please cite

> Vandenbulcke S, Vanderaa C, Crook O, Martens L, Clement L.
msqrob2TMT: robust linear mixed models for inferring differential
abundant proteins in labelled experiments with arbitrarily complex
design. bioRxiv. Published online March 29, 2024:2024.03.29.587218.
doi:10.1101/2024.03.29.587218

**TODO**: populate the citation section and 

## Session Info

The following packages have been used to generate this document.

```{r, echo=FALSE}
devtools::load_all()
```

```{r}
sessionInfo()
```

